---
title: "experiment_ebpm_bg_kos"
author: "zihao12"
date: "2020-04-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
I want to try `ebpm-background` model on the fitted model on kos data: https://zihao12.github.io/ebpmf_demo/applications_kos.html

```{r}
rm(list = ls())
library(pheatmap)
source("code/misc.R")
set.seed(123)

data_dir = "~/Desktop/data/text"
dict_name = "vocab.kos.txt"
fitted_file = "data/experiment_ebpm_bg_kos2_data.Rds"

dict = read.csv(sprintf("%s/%s", data_dir, dict_name), header = FALSE)
dict = as.vector(dict[,1])
```

I construct a poisson means sub-problem this way
\begin{align}
& Y_{jk} = \sum_i Z_{ijk}\\
& Z_{ijk} = X_{ij} \zeta_{ijk}\\
& \zeta_{ijk} = \frac{l_{ik} f_{jk}}{\sum_k l_{ik} f_{jk}}\\
\end{align}


Then fit `ebpm_background` model on $Y_{jk}$
```{r}
fitted = readRDS(fitted_file)
Y = fitted$Y
model = fitted$fit
plot(model$progress)
K = ncol(Y)
p = nrow(Y)
s_k = colSums(fitted$L)

s_k

V_bg = model$posterior$mean
mu_bg = model$mu
Lam_bg = mu_bg * V_bg

Lam_mle = t(t(Y)/s_k)
mu_median = apply(Lam_mle, 1, median)
```


## $\mu_j$
* even the largest background frequency is too small (should we do scaling?)\

* majority background frequencies should probably just be 0...
```{r}
hist(mu_median, breaks = 100, xlim = c(0,100))
## what percentage of words have background (use median of Lam_mle) 0
sum(mu_median == 0)/nrow(Y)
## max background
max(mu_median)

plot(mu_median, mu_bg, pch = 20)
abline(0,1,lwd=3, col = "green")
```

## $\lambda_{jk}$
* Not too much different from MLE result. \

```{r }
plot(Lam_mle, Lam_bg)
## look at those with bigger background freq
mask <- (mu_median > 20)
plot(Lam_mle[mask,], Lam_bg[mask,])
```

## look at some $\lambda_{jK}$
```{r fig.width=14, fig.height=14}

lf = poisson2multinom(F = fitted$F, L = fitted$L)

par(mfrow = c(5,4))
for(k in 1:K){
  j = which.max(lf$F[,k])
  plot(Lam_mle[j,], xlab = "topic index", ylab = "f_jk",
       main = sprintf("f_jK for `%s` in topic %d", dict[j], k))
  points(Lam_bg[j,], col = "blue")
}

```



