---
title: "compare `ebpm` with `gauss_gh`"
author: "zihao12"
date: "2019-12-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
I want to compare `ebpm` with the algorithm `Gauss-HG` propsed in paper  [Bayesian inference on quasi-sparse count data 
](https://academic.oup.com/biomet/article/103/4/971/2659041). Below I first copy from their analysis http://dattahub.github.io/GHstancodes , then compare `ebpm` with theirs. 

```{r warning=FALSE, message=FALSE}
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
library(ggplot2)
theme_set(theme_bw())
library(plyr)
library(dplyr)
library(reshape2)
```

## `Gauss-HG` algorithm
```{r cache=TRUE, warning=F, message=FALSE}
# setup Stan Gauss-HG sampler
{
  library(plyr)
  library(rstan)
  library(parallel)
  library(rbenchmark)
  
  #set_cppo("fast")
  stan.gh.code = "
  data{
  int<lower=0> J;
  int<lower=0> Y[J];
  real<lower=0> alpha;
  real<lower=0> a;
  real<lower=0> b;
  real<lower=0> gamma;
  real<lower=0> phi;
  }
  parameters{
  real<lower=0,upper=1> kappa[J];
  real<lower=0> theta[J];
  }
  model{
  for(i in 1:J) {
  increment_log_prob((a-1)*log(kappa[i])+(b-1)*log(1-kappa[i])-gamma*log(1-phi*kappa[i]));
  theta[i] ~ gamma(a, kappa[i]/(1-kappa[i]));
  Y[i] ~ poisson(theta[i]);
  }
  }
  "
  stan.gh.fit = stan_model(model_code=stan.gh.code, model_name="GH")
}
```

## simulation
```{r warning=F}
stan.iters = 10000
n.chains = 2
seed.val = 786
set.seed(seed.val)

n = 200; w = 0.9
y = rep(0,n); idx = rep(1,n)
lambdasparse = rep(0,n)
for (i in 1:n)
{
  if(i<=round(n*w)){
    lambdasparse[i]<-0.1
    idx[i] <- 0}
  else {lambdasparse[i] <-10}}

y = rpois(n,lambdasparse); 
gamma = mean(kmeans(y,centers=2)$centers)
alpha = 0.01
a = 0.5; b = 0.5
gh.data = list('J'=n,'Y'=y, 'alpha' = alpha,'a' = a, 'b' = b, 'gamma' = gamma, 'phi' = 0.99)
```

## fit with `Gauss-HG`
```{r cache=TRUE}
{
  gh.res = sampling(stan.gh.fit, 
                     data = gh.data, 
                     iter = stan.iters,
                     warmup = floor(stan.iters/2),
                     thin = 2,
                     pars = c('kappa','theta'),
                     init = 0,
                     seed = seed.val, 
                     chains = 1)
  
  gh.theta.smpls = extract(gh.res, pars=c('theta'), permuted=TRUE)[[1]]
  gh.kappa.smpls = extract(gh.res, pars=c('kappa'), permuted=TRUE)[[1]]
  gh.theta.mean = apply(gh.theta.smpls,2,mean)
  gh.kappa.mean = apply(gh.kappa.smpls,2,mean)
  
  gh.sample.data = melt(extract(gh.res, permuted=TRUE))
  colnames(gh.sample.data) = c("iteration", "component", "value", "variable")
  
  gh.sample.data= gh.sample.data %>%
    filter(variable %in% c("theta","kappa")) 

  gh.sample.data.2 = gh.sample.data %>% group_by(component, variable) %>%
  summarise(upper = quantile(value, prob=0.975), 
            lower = quantile(value, prob=0.225),
            middle = mean(value))
}
```



## fit with `ebpm`
```{r warning=F}
library(ebpm)
fit_ebpm_gammamix = ebpm_gamma_mixture_single_scale(x = y, s = 1)
fit_ebpm_expmix = ebpm_exponential_mixture(x = y, s = 1)
fit_ebpm_pg = ebpm_point_gamma(x = y, s = 1)
fit_ebpm_tg = ebpm_two_gamma(x = y, s = 1, rel_tol = 1e-8)
```


```{r}
fit_df = data.frame(
  data = y,
  lam_true = lambdasparse,
  gh = gh.theta.mean,
  ebpm_pg = fit_ebpm_pg$posterior$mean,
  ebpm_tg = fit_ebpm_tg$posterior$mean,
  ebpm_expmix = fit_ebpm_expmix$posterior$mean,
  ebpm_gammamix = fit_ebpm_gammamix$posterior$mean
)

ggplot(data = fit_df)+
  geom_point(aes(x = data, y = lam_true, color = "lam_true"))+
  geom_point(aes(x = data, y = gh, color = "gauss-hg"))+
  geom_point(aes(x = data, y = ebpm_pg, color = "ebpm_point_gamma"))+
  geom_point(aes(x = data, y = ebpm_tg, color = "ebpm_two_gamma"))
```

Take a closer look at those quasi-zeros (counts that comes from small lambda)
```{r}
fit_df_small = fit_df[fit_df$lam_true < 1, ]
ggplot(data = fit_df_small)+
  geom_point(aes(x = data, y = lam_true, color = "lam_true"))+
  geom_point(aes(x = data, y = gh, color = "gauss_hg"))+
  geom_point(aes(x = data, y = ebpm_pg, color = "ebpm_point_gamma"))+
  geom_point(aes(x = data, y = ebpm_tg, color = "ebpm_two_gamma"))
```


Below I show the divergence between estimation and truth (Root Mean Squared Error, Kullback–Leibler divergence , Jensen-Shannon)
```{r}
rmse <- function(true, est){
  return(sqrt(mean((true - est)^2)))
}
KL <- function(true,est){
  sum(ifelse(true==0,0,true * log(true/est)) + est - true)
}
JS  <- function(true,est){
  0.5*(KL(true, est) + KL(est, true))
}
RMSEs = c(rmse(lambdasparse, gh.theta.mean), rmse(lambdasparse, fit_ebpm_gammamix$posterior$mean), 
                     rmse(lambdasparse, fit_ebpm_expmix$posterior$mean), 
                     rmse(lambdasparse, fit_ebpm_pg$posterior$mean),
                     rmse(lambdasparse,fit_ebpm_tg$posterior$mean))

KLs = c(KL(lambdasparse, gh.theta.mean), KL(lambdasparse, fit_ebpm_gammamix$posterior$mean), 
                     KL(lambdasparse, fit_ebpm_expmix$posterior$mean), 
                     KL(lambdasparse, fit_ebpm_pg$posterior$mean),
                     KL(lambdasparse,fit_ebpm_tg$posterior$mean))

JSs = c(JS(lambdasparse, gh.theta.mean), rmse(lambdasparse, fit_ebpm_gammamix$posterior$mean), 
                     JS(lambdasparse, fit_ebpm_expmix$posterior$mean), 
                     JS(lambdasparse, fit_ebpm_pg$posterior$mean),
                     JS(lambdasparse,fit_ebpm_tg$posterior$mean))
data.frame(RMSE = RMSEs, KL = KLs, JS = JSs, row.names = c("guass-hg", "ebpm_gammamix", "ebpm_expmix", "ebpm_point_gamma", "ebpm_two_gamma"))

```

### Comment:
1. `GH` shrinks too much. Type-I error seems indeed pretty small, as proved in the paper. The expense is the very bad estimates for bigger counts. Maybe need to choose different hyperparameters.  

2. `ebpm_point_gamma` fails for those "quasi-sparse" counts, the point-mass at 0 for prior won't affect their posteriors. They also affect the estimation for larger counts.  

3. `ebpm_two_gamma` performs the best on average. It slightly overestimates those "quasi-sparse" counts, but is very close to truth overall. 

4. `ebpm_expmix` and `ebpm_gammamix` does not do well. Only two prior components are effectively used, and certainly not as well-chosen as `gamma_two_gamma`. (didn't show in the plot)


