---
title: "current_progress_ebpmf_comparison"
author: "zihao12"
date: "2020-02-25"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
I show ther current [progress](https://zihao12.github.io/ebpmf_demo/current_progress_ebpmf) here. Now I want to compare the results with existing algorithms. Here I want to show `nsNMF`


## Non-smooth NMF. 
Uses a modified version of Lee and Seung’s multiplicative updates for Kullback-Leibler divergence to fit a extension of the standard NMF model. It is meant to give sparser results.

Reference: [Pascual-Montano2006](https://ieeexplore.ieee.org/document/1580485)

The tuning parameter $\theta \in [0, 1]$ corresponds to the non-smoothness or sparseness of the solution. When $\theta = 0$, its solution should be the same as MLE.

## Result
We can get sparse solution when using big $\theta$, however the resulting $\hat{\lambda}$ is further from truth. 

```{r}
rm(list = ls())
library(NMF)
set.seed(123)
```

```{r}
KL <- function(true,est){
  sum(ifelse(true==0,0,true * log(true/est)) + est - true)
}

JS  <- function(true,est){
  0.5*(KL(true, est) + KL(est, true))
}

RMSE <- function(true, est){
  sqrt(mean((true - est)^2))
}

log_lik <- function(X, lam){
  return(sum(dpois(x = X, lambda = lam , log = T)))
}

show_lf <- function(fit, name, l = TRUE, f = FALSE){
  L = fit@W 
  F = t(fit@H)
  k = ncol(L)
  if(l){
    par(mfrow = c(2,2))
    for(i in 1:k){
      plot(L[,i],main=sprintf("theta = %s: loading %d",name, i), ylab = "loading", pch = 16)
    }
  }
  if(f){
    par(mfrow = c(2,2))
    for(i in 1:k){
      plot(F[,i],main=sprintf("%s: factor %d",name, i), ylab = "factor", pch = 16)
    }
  }
}
```

Load results from [progress](https://zihao12.github.io/ebpmf_demo/current_progress_ebpmf)
```{r}
results = readRDS("data/out_current_progress_ebpmf.Rds")
X = results$X
L = results$L
F = results$F
lambda = L %*% t(F)
fit_tg = results$fit_tg
fit_lee = results$fit_lee
k = ncol(L)
```

## Fit with different $\theta$
```{r cache=TRUE, autodep=TRUE}
fits <- c()
rmse <- c()
kl <- c()
js <- c()
ll <- c()
thetas = c(0, 0.2, 0.4, 0.6, 0.8)
#thetas = c(0.2, 0.4)
for(theta in thetas){
  fit_ = fit_nsNMF = nmf(X,k,method = 'nsNMF', theta = theta)@fit
  lam_ = fit_@W %*% fit_@H
  rmse <- c(rmse, RMSE(lambda, lam_))
  kl <- c(kl, KL(lambda, lam_))
  js <- c(js, JS(lambda, lam_))
  ll <- c(ll, log_lik(X, lam_))
  fits <- c(fits, fit_)
}
names(fits) <- thetas
#fit_nsNMF = nmf(X,k,method = 'nsNMF', theta = 0.9)
```

## look at loadings
```{r}
for(name in names(fits)){
  print(sprintf("theta = %s", name))
  show_lf(fits[[name]], name)
}
```
Although we can get some smooth solution by using large $\theta$, it seems to give bad estimates of the non-zero parts. Let's see if the smoothness help recover the truth.

## look at the divergence of $\hat{\lambda}$ from true $\lambda$
```{r}
par(mfrow = c(2,2))
plot(thetas, rmse, xlab = "theta", ylab = "rmse", log = "y", pch = 16)
plot(thetas, kl, xlab = "theta", ylab = "kl", log = "y" ,pch = 16)
plot(thetas, js, xlab = "theta", ylab = "js", log = "y", pch = 16)
plot(thetas, ll, xlab = "theta", ylab = "loglikelihood", pch = 16)
```
So the imposed sparseness does not help recover the truth.
























<!-- ## Sparse NMF -->

<!-- Alternating Least Square (ALS) approach. It is meant to be very fast com- pared to other approaches. -->

<!-- Reference: ![KimH2007](https://www.ncbi.nlm.nih.gov/pubmed/17483501) -->

<!-- ```{r} -->

<!-- fit_snmfl = nmf(X,k,method = 'snmf/l',beta=20) -->

<!-- fit_snmfr = nmf(X,k,method = 'snmf/r',beta=1) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- par(mfrow=c(2,2)) -->
<!-- for(i in 1:k){ -->
<!--     plot(fit_nsNMF@fit@W[,i],main=paste0("nsNMF: estimated loadings ",i)) -->
<!-- } -->

<!-- par(mfrow=c(2,2)) -->
<!-- for(i in 1:k){ -->
<!--     plot(fit_snmfl@fit@W[,i],main=paste0("snmfl: estimated loadings ",i)) -->
<!-- } -->

<!-- par(mfrow=c(2,2)) -->
<!-- for(i in 1:k){ -->
<!--     plot(fit_snmfr@fit@W[,i],main=paste0("snmfr: estimated loadings ",i)) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r} -->
<!-- par(mfrow=c(2,2)) -->
<!-- for(i in 1:k){ -->
<!--     plot(fit_nsNMF@fit@H[i,],main=paste0("nsNMF: estimated factors ",i)) -->
<!-- } -->

<!-- par(mfrow=c(2,2)) -->
<!-- for(i in 1:k){ -->
<!--     plot(fit_snmfl@fit@H[i,],main=paste0("snmfl: estimated factors ",i)) -->
<!-- } -->

<!-- par(mfrow=c(2,2)) -->
<!-- for(i in 1:k){ -->
<!--     plot(fit_snmfr@fit@H[i,],main=paste0("snmfr: estimated factors ",i)) -->
<!-- } -->
<!-- ``` -->






