---
title: "ebpm_two_gamma_debug1"
author: "zihao12"
date: "2020-01-16"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
There were issues with `ebpm_two_gamma` (when used in `ebpmf`). I reproduce the problem below. 

```{r}
library(ebpm)
data = readRDS("../ebpm/data/ebpmf_two_gamma_issue2.Rds")
x = data$x
s = data$s
hist(x)
print(s)

print("the following are the possible ll from ebpm (initialized with K-means)")
replicate(20,ebpm_two_gamma(x = x, s = s)$log_likelihood)
```

### look at results from `ebpm_point_gamma`
For `ebpm_point_gamma`, find out why it can have positive ll. 
```{r}
stop = FALSE
while(!stop){
  fit_ = ebpm_two_gamma(x, s)
  ll = fit_$log_likelihood
  if(ll > 0){
    stop = TRUE
    print(sprintf("ll = %f", ll))
    print("fitted_g")
    print(fit_$fitted_g)
  }
}
```

well, when `a` and `b` gets too big, the computation of NB likelihood becomes inaccurate (even positive). 
```{r}
## (in ebpm package) the obj function
tg_nlm_fn <- function(par, x, s){
  ## d1 = NB(x, a1, b1/(b1+s)); d2 = NB(x, a2, b2/(b2+s))
  ## return - log(pi0 * d1 + (1-pi0) * d2)
  #browser()
  pi0 = 1/(1+ exp(-par[1]))
  a1 = exp(par[2])
  b1  =  exp(par[3])
  a2 = exp(par[4])
  b2  =  exp(par[5])
  d1_log <- dnbinom_cts_log_vec(x, a1, b1/(b1+s))
  d2_log <- dnbinom_cts_log_vec(x, a2, b2/(b2+s))
  out = sum( log( pi0*exp(d1_log)  + (1-pi0)*exp(d2_log) ) )
  return(-out)
}


## (in ebpm package) function that computes NB loglikelihood
dnbinom_cts_log_vec <- function(x, a, prob){
  tmp = x*log(1-prob)
  tmp[x == 0] = 0 ## R says 0*-Inf = NaN
  return(a*log(prob) + tmp + lgamma(x+a) - lgamma(x+1) - lgamma(a))
}

## use jensen inequality:
## sum( log( pi0*exp(d1_log)  + (1-pi0)*exp(d2_log) ) ) >=  sum ( pi0*d1_log + (1-pi0)*d2_log ) = pi0 * sum(d1_log) + (1-pi0)*sum(d2_log)
## below I show sum(d1_log) & sum(d2_log)

pi0 = fit_$fitted_g$pi0

## likelihood for mixture 1
a = fit_$fitted_g$shape1
b = 1/fit_$fitted_g$scale1
prob = b/(b + s)
sum_d1_log = sum(dnbinom_cts_log_vec(x, a, prob))
sum_d1_log

a = fit_$fitted_g$shape2
b = 1/fit_$fitted_g$scale2
prob = b/(b + s)
sum_d2_log = sum(dnbinom_cts_log_vec(x, a, prob))
sum_d2_log

## the lower bound for loglikelihood p(x | s, pi0, a1, a2, b1, b2)
pi0 * sum_d1_log + (1 - pi0) * sum_d2_log
```
So the problem is our `dnbinom_cts_log_vec` function can't handle extremely large `a` and `b`. Think the problem as something to do with `lgamma` for big numbers. See `numerical_lgamma.html`. \

Also, why it goes to infinity during the optimization?



