---
title: "ebpm_debug"
author: "zihao12"
date: "2019-09-23"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Goal
My code does not always get $RMSE(posterior, oracle) <  RMSE(MLE, oracle)$. Therefore  I  just use my model to generate posterior mean, which according to my derivation should be $$\sum_k \pi_k \frac{x_i + a_k}{s_i + b_k}$$
where $\pi_k$ is known. \\
See detail in  https://www.overleaf.com/project/5bd084d90a33772e7a7f99a2

```{r}
library(mixsqp)
library(gtools)

```


## EBPM-exponential-mixture
* For now, I use mixture of exponential as prior family.
* Under the exponential case, we discuss how to select the range of  the grid (of $\mu$, exponential mean). For convenience we use $exp(\mu)$ to denote exponential distribution with mean $\mu$:\
The goal is: for each observation, we want to include the range of $\lambda$ "of  interest" (i.e. $log \ p(x | \lambda)$ is  close to that of the MLE, like within $log(0.1)$). \
If $x = 0$, $\ell(\lambda) = - \lambda s$, in order to  have a good likelihood, we want the model to be able to choose  $\lambda \sim o(\frac{1}{s})$. Therefore, we want the smallest $\mu$ to be  in the order of $o(\frac{1}{s})$ if there  is 0 count.\
If $x > 0$, the MLE would be $\frac{x}{s}$, and $\lambda$ too small would have bad likelihood. So we want the biggest $\mu$ to be of order $O(max(\frac{x}{s}))$

## experiment setup
* I simulate $\lambda_i \sim \sum_k \pi_k exp(b_k), k = 1, ...,  50$, where $b_k$ is the exponential rate, for $i = 1, ..., 4000$. 
* Then I fit  EBPM with mixture of exponential as prior (our model knows the grid for $b_k$). I compare $\ell(\pi)$, which should be better than oracle; I also compare $\ell(\lambda)$ (it is not clear whether the posterior  will be better than oracle, but still  we can use it  to see how good the model is). 
*  Then  I fit with the same model, without knowing the grid for $b_k$. 


```{r}
## ===========================================================================
## ==========================ebpm_exponential_mixture=========================
## ===========================================================================

## debug ebpm_exponential_mixture problem
ebpm_exponential_mixture_debug <- function(x,s, grid, pi){
  a = grid$a
  b = grid$b
  cpm = outer(x,a,  "+")/outer(s, b, "+")
  lam_pm = cpm %*% pi
  ll_lam = sum(dpois(x, s*lam_pm, log = T))
  return(list(pi = pi, lam_pm = lam_pm, ll_lam = ll_lam,grid = grid))
}

## ===========================================================================
## ==========================experiment setup=================================
## ===========================================================================
## sample from mixture of gamm distribution
sim_mgamma <- function(a,b,pi){
  idx = which(rmultinom(1,1,pi) == 1)
  return(rgamma(1, shape = a[idx], rate =  b[idx]))
}

## simulate a poisson mean problem
simulate_pm  <-  function(n, d, seed = 123){
  set.seed(seed)
  ## simulate grid
  a = replicate(d,1)
  b = 10*runif(d)
  grid  = list(a = a, b = b)
  pi <- matrix(rdirichlet(1,rep(1/d, d)), ncol = 1)
  lam_true = replicate(n, sim_mgamma(a,b,pi))
  s = replicate(length(lam_true), 1)
  #s = 2*runif(length(lam_true))
  x  = rpois(length(lam_true),s*lam_true)
  ll_lam = sum(dpois(x, s*lam_true, log = T))
  return(list(x =  x, s = s, lam_true = lam_true, pi = pi, grid = grid, ll_lam = ll_lam))
}

rmse <- function(x,y){
  return(sqrt(mean(x-y)^2))
}
```

```{r}
## test functions  above
main <- function(n=1000, d = 50){
  sim = simulate_pm(n, d)
  x = sim$x 
  s = sim$s
  lam_true = sim$lam_true
  
  fit = ebpm_exponential_mixture_debug(x, s,sim$grid, sim$pi)

  print(sprintf("%d data points and %d grid points", length(sim$x),length(fit$grid$b)))
  print("RMSE with lam_oracle:")
  print(sprintf("mle : %f", rmse(x/s, sim$lam_true)))
  print(sprintf("pos : %f", rmse(fit$lam_pm, sim$lam_true)))
}
```


## fit 
```{r}
for(n in 10^(seq(1,6,1))){
  main(n)
}
```

