---
title: "Experiment_ebpmf_rank1"
author: "zihao12"
date: "2019-10-26"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Goal

* I fit `ebpmf` on `rank-1` dataset with $K = 2$, and see if we are able to the true $K$ from the result.                                     

* From the [Comparison between `scd` and `lee`](https://zihao12.github.io/ebpmf_demo/Compare_scd_lee.html), I think the result would be very different if we initialize the algortihm with `scd` or `lee` (from `NNLM::nnmf`). Also, it would be helpful to see how our algorithm changes from the initialization. 

## Setup
Data generating model: here the mean structure is rank-1. 
$$
\begin{align}
  & \Lambda_{ij} = l f_j\\
  & X_{ij} \sim Pois(\Lambda_{ij})
\end{align}
$$
There are two scenarios:\

* $l^1 \approx l, l^2 \approx 0$.\

* $l^1 \approx cl, l^2 \approx (1-c)l$. \

We prefer the first one for our application.

## Result
* `ebpmf` is influenced a lot by the initialization. It is not surprising as the initialization heavily affects how we partition $X_{ij} = \sum_k Z_{ijk}$. \

* `ebpmf` seems to be turning the first scenario to the second scenario gradually (compare `lee` and  `ebpmf` initialized from `lee`), which is alarming. 


```{r}
rm(list = ls())
devtools::load_all("../ebpmf")
library(ebpmf)
```


```{r}
sim_pois_rank1 <- function(n, p, seed = 123){
  set.seed(seed)
  L = matrix(replicate(n, 1), ncol = 1)
  F = matrix(sample(seq(1,1000,length.out = p)), ncol = 1)
  Lam = L %*% t(F)
  X = matrix(rpois(n*p, Lam), nrow = n)
  Y = matrix(rpois(n*p, Lam), nrow = n)
  ll_train = sum(dpois(X, Lam, log = T))
  ll_val = sum(dpois(Y, Lam, log = T))
  return(list(X = X, Y = Y, L = L, F = F, Lam = Lam, ll_train = ll_train, ll_val  = ll_val))
}

# Scale each column of A so that the entries in each column sum to 1;
# i.e., colSums(scale.cols(A)) should return a vector of ones.
scale.cols <- function (A)
  apply(A,2,function (x) x/sum(x))

# Convert the parameters (factors & loadings) for the Poisson model to
# the factors and loadings for the multinomial model. The return value
# "s" gives the Poisson rates for generating the "document" sizes.
poisson2multinom <- function (F, L) {
  L <- t(t(L) * colSums(F))
  s <- rowSums(L)
  L <- L / s
  F <- scale.cols(F)
  return(list(F = F,L = L,s = s))
}

show_loadings <- function(L, title = "hist for two loadings"){
  L_df = data.frame(L)
  colnames(L_df) = c("loading1", "loading2")
  hist(L_df$loading1, col = "red", xlim=c(0, 1), xlab = "loading proportion", main = title)
  hist(L_df$loading2, col = "blue", add = T)
}
```



Simulate data
```{r}
n = 50
p = 100
sim = sim_pois_rank1(n, p)
```

## Fit with `ebpmf` initialized from `nnmf` using  `lee` and `scd`
```{r warning=F, message=F}
K = 2
init_lee = NNLM::nnmf(A = sim$X, k = K, loss = "mkl", method = "lee", max.iter = 1000)
init_scd = NNLM::nnmf(A = sim$X, k = K, loss = "mkl", method = "scd", max.iter = 1000)

## ebpmf_point_gamma init with lee
fit_init_lee = ebpmf::ebpmf_point_gamma(sim$X, K = K, maxiter.out = 100, 
                                        qg = initialize_qg_from_LF(L0 = init_lee$W, F0 = t(init_lee$H)))
fit_init_scd = ebpmf::ebpmf_point_gamma(sim$X, K = K, maxiter.out = 100, 
                                        qg = initialize_qg_from_LF(L0 = init_scd$W, F0 = t(init_scd$H)))
```

Let's see how the loadings change for `ebpmf` initialized with scd
```{r}
p2m_res = poisson2multinom(F  = t(init_scd$H), L = init_scd$W)
show_loadings(p2m_res$L, title = "scd")

p2m_res = poisson2multinom(F  = fit_init_scd$qg$qfs_mean, L = fit_init_scd$qg$qls_mean)
show_loadings(p2m_res$L, title = "ebpmf init from scd")
```


Let's see how the loadings change for `ebpmf` initialized with scd
```{r}
p2m_res = poisson2multinom(F  = t(init_lee$H), L = init_lee$W)
show_loadings(p2m_res$L, title = "lee")

p2m_res = poisson2multinom(F  = fit_init_lee$qg$qfs_mean, L = fit_init_lee$qg$qls_mean)
show_loadings(p2m_res$L, title = "ebpmf init from lee")
```

Note that `ebpmf` is turning  the first scenario to the second!!


