---
title: "experiment_nb_means5"
author: "zihao12"
date: "2020-04-09"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
Investigate the model (suppress index $j$ as each row is independent)

\begin{align}
& y_k \sim Po(s_k \lambda_k)\\
& \lambda_k = \mu Ga(1/\phi_k, 1\phi_k)
\end{align}

We want to focus on the case of a "sparse" sequence: most of $y$ is close to a constant, and few deviates a lot. 

## Conclusion
* In these very simple examples, EB does not have much shrinkage effect on outlier counts \

* We can get meaningful $\phi$ and "right" $\mu$ from EB approach, if initialized properly


```{r}
rm(list = ls())
source("script/nb_means.R")
set.seed(123)
maxiter = 100
verbose = FALSE 
```


## on an ariticial sequence

* I have the $\mu$ around $10$ for the majority counts. I append shorter sequence of downward/upperward deviations. 

* The result is not much different from MLE (don't have much shrinkage effect)

* The $\phi$ and $\mu$ makes sense. 

### using descend method
```{r message=FALSE, warning=FALSE}
y = c(rpois(n = 20, lambda = 2),
      rpois(n = 100, lambda = 20), 
      rpois(n = 20, lambda = 100))

## look at sequence y
y

Y1 = t(matrix(replicate(2, y), ncol = 2))
s_ = replicate(length(y), 1)

mle1 = t(t(Y1)/s_)


fit1 <- nb_means(Y = Y1, s = s_, mu = NULL, A = NULL, maxiter = maxiter, verbose = verbose,
                              control = list(method = "descend", gradient = TRUE, hessian = FALSE), seed = 1234)
phi1 = 1/fit1$fitted_g$A
mu1 = fit1$fitted_g$mu
pm1 = fit1$posterior$mean

plot(mle1[1,], col = "red", pch = 20)
points(pm1[1,], col = "blue", pch = 18)
abline(h = mu1[1], col = "black")
legend("topleft", legend=c("mle", "eb post-mean", "mu"),
       col=c("red","blue", "black"), lty=c(1,1,1), cex=0.8)

plot(phi1[1,], log = "y")
mu1[1]
```


### using grid search
```{r message=FALSE, warning=FALSE}
fit1 <- nb_means(Y = Y1, s = s_, mu = NULL, A = NULL, maxiter = maxiter, verbose = verbose,
                              control = list(method = "grid", gradient = TRUE, hessian = FALSE), seed = 1234)
phi1 = 1/fit1$fitted_g$A
mu1 = fit1$fitted_g$mu
pm1 = fit1$posterior$mean

plot(mle1[1,], col = "red", pch = 20)
points(pm1[1,], col = "blue", pch = 18)
abline(h = mu1[1], col = "black")
legend("topleft", legend=c("mle", "eb post-mean", "mu"),
       col=c("red","blue", "black"), lty=c(1,1,1), cex=0.8)

plot(phi1[1,],log = "y")
mu1[1]
```


## on fitted $F$ from single-cell data
```{r}
load("data/cytokines_data_fit.RData")
fit.F[fit.F < 1e-14] = 0 ## they should be 0 (only set to 1e-15 for numerical issue)

n = nrow(fit.F)
idx = 1: n
gene_sub_names = rownames(fit.F)[idx]
gene_ranked = readRDS("./data/genes_ranked.RDS")
id_by_std = order(gene_ranked[gene_sub_names,"variance.standardized"], decreasing = TRUE)

## look at the most variable genes (not reliable)
fit.F[id_by_std[1:5],]

## look at the least variable genes
fit.F[id_by_std[(n-5):n],]

## look at s_k
s_k
```


```{r}
s_ = s_k
K = length(s_)
p = 5
Lambda1 = fit.F[id_by_std[1:p],]
Y1 = matrix(rpois(n = p * K, lambda = Lambda1 %*% diag(s_)), ncol = K)

## look at the sequence
Y1[1,]/s_

## MLE 
mle1 = t(t(Y1)/s_)

## EB fit
fit1 <- nb_means(Y = Y1, s = s_, mu = NULL, A = NULL, maxiter = maxiter, verbose = verbose,
                              control = list(method = "grid", gradient = TRUE, hessian = FALSE), seed = 1234)
phi1 = 1/fit1$fitted_g$A
mu1 = fit1$fitted_g$mu
pm1 = fit1$posterior$mean

plot(Lambda1[1,], col = "green", pch = 18)
points(mle1[1,], col = "red")
points(pm1[1,], col = "blue")
abline(h = mu1[1], col = "black")
legend("topleft", legend=c("truth","mle", "eb post-mean", "mu"),
       col=c("green", "red","blue", "black"), lty=c(1, 1,1,1), cex=0.8)

mu1[1]
plot(phi1[1,])
```
* Obviously, MLE is good enough here. 

* Although posterior mean looks fine, EB seems to get the wrong structure (local optimum?). Let's re-initialize it (with the "right" $\mu$)


```{r}
fit1 <- nb_means(Y = Y1, s = s_, mu = replicate(5, 0), A = NULL, maxiter = maxiter, verbose = verbose,
                              control = list(method = "grid", gradient = TRUE, hessian = FALSE), seed = 1234)

phi1 = 1/fit1$fitted_g$A
mu1 = fit1$fitted_g$mu
pm1 = fit1$posterior$mean

plot(Lambda1[1,], col = "green", pch = 18)
points(mle1[1,], col = "red")
points(pm1[1,], col = "blue")
abline(h = mu1[1], col = "black")
legend("topleft", legend=c("truth","mle", "eb post-mean", "mu"),
       col=c("green", "red","blue", "black"), lty=c(1, 1,1,1), cex=0.8)

mu1[1]

plot(phi1[1,])
```


