---
title: "f21_deriv3"
author: "zihao12"
date: "2020-02-09"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
I want to find a good way of computing $_2 F_{1}(a, b ; c ; z)$ and $_2 F_{1}^{(0,0,1,0)}(a, b ; c ; z)$ ($b$ is a vector). 

## Methods and Results
* For computing $_2 F_{1}$, I use `BMS::f21hyper` (vectorized), `hypergeo::hypergeo` and `gsl::hyperg_2F1` (use transformations for some values of $z$, which I name to be `Gauss2F1b`; this methods is from [a question in stack overflow](https://stats.stackexchange.com/questions/33451/computation-of-hypergeometric-function-in-r)). `Gauss2F1b` is much faster (I didn't try to figure out why). \

* For computing  $_2 F_{1}^{(0,0,1,0)}(a, b ; c ; z)$, I use numerical differentiation. I only tried `numDeriv::grad`, with `Richardson`. I will explore more methods if the current methods are not fast enough. ([mathworld](http://mathworld.wolfram.com/) provides two formula for computing it, but both seem complicated to implement)


```{r}
rm(list = ls())
library(BMS)
library(numDeriv)
library(gsl) ## need to install GSL to computer: https://astrointro.wordpress.com/2017/05/17/installing-gnu-scientific-library-gsl-in-ubuntu-16-04-and-compiling-codes/ &&  https://blog.csdn.net/qtyl1988/article/details/25636555 

f21.my <- function(A, B, C, z){
  if(z < 1 && z > -1){return(f21hyper(A,B,C,z))}
  return(+Inf) ## not sure for z < -1
}

f21.my.vector <- Vectorize(f21.my)
hypergeo.vector = Vectorize(hypergeo::hypergeo) ## not sure if vectorization is necessary here


## I got a very good method from stack overflow: https://stats.stackexchange.com/questions/33451/computation-of-hypergeometric-function-in-r
Gauss2F1 <- function(A,B,C,z){
    if(z>=0 & z<1){
        hyperg_2F1(A,B,C,z)
    }else{
            hyperg_2F1(C-A,B,C,1-1/(1-z))/(1-z)^B
        }
}
```


```{r}
f21deriv3.exp <- function(nz = 2000, nnz = 1000, seed = 123){
  n = nz + nnz
  print(sprintf("problem size is %d", n))
  set.seed(seed)
  ## simulate data
  x = c(replicate(nz, 0), rpois(n = nnz, lambda = 10))
  ## use default parameters for g
  a = b = alpha = 0.5
  gam = mean(kmeans(x = x, centers = 2)$centers)
  phi = min(0.9, sum(x > 0)/sum(x == 0))
  g = list(a = a, b = b, alpha = alpha, gam = gam, phi = phi)
  
  ## parameter for posterior of psi
  a = g$a + g$alpha
  b = g$b + x
  gam = g$gam
  z = - (1 - g$phi)
  
  
  ## use runtime for computing beta as benchmark
  print("compute beta as benchmark")
  start = proc.time()
  test0 = beta(a = a, b = b)
  runtime = proc.time() - start
  print(sprintf("compute beta: %s seconds", runtime[[3]]))
  
  
  
  ################################## compare runtime for computing f21
  print("################")
  print("compute f21 with different methods")
  ## f21hyper
  start = proc.time()
  test1 = f21.my.vector(A = gam, B = a, C = a + b, z = z)
  runtime = proc.time() - start
  print(sprintf("compute f21 with f21hyper: %s seconds", runtime[[3]]))
  
  ## hypergeo
  start = proc.time()
  test2 = hypergeo::hypergeo(A = gam, B = a, C = a + b, z = z)
  runtime = proc.time() - start
  print(sprintf("compute f21 with hypergeo: %s seconds", runtime[[3]]))
  
  ## 
  start = proc.time()
  test3 = Gauss2F1(A = gam, B = a, C = a + b, z = z)
  runtime = proc.time() - start
  print(sprintf("compute f21 with Gauss2F1: %s seconds", runtime[[3]]))
  print(sprintf("max difference between f21hyper and Gauss2F1 is %20f", max(abs(test1 - test3))))
  
  
  ## compute derivative
  print("################")
  print("now we use Gauss2F1 to compute gradient")

  ## compute derivative
  start = proc.time()
  grad(Gauss2F1, x = a + b, A = gam, B = a, z = z, method = "Richardson")
  runtime = proc.time() - start
  print(sprintf("Richardson's extrapolation method: %s seconds", runtime[[3]]))
}
```


```{r}
f21deriv3.exp(nz = 20000, nnz = 10000)
```



