---
title: "fit_cytokines_data"
author: "zihao12"
date: "2020-04-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## experiment setup

* The idea is to simulate realistic data from fitted `F`, `s` on `cytokines` data (provided by Yusha). (`F` becomes $\lambda$ in the model below)\

* I simulate and fit data using the model below:
\begin{align}
  & Y_{jk} \sim Po(s_k \lambda_{jk})\\
  & \lambda_{jk} = \mu_j v_{jk}\\
  & v_{jk} \sim Ga(1/\phi_{jk}, 1/\phi_{jk})
\end{align}
I use the $\lambda_{jk}$ from fitted $F$ to simulate data; obtain $\mu_j, \phi_{jk}$  and posterior mean of $\lambda$ from the fit. 


```{r}
rm(list = ls())
load("data/cytokines_data_fit.RData")
source("script/nb_means.R")
set.seed(123)
fit.F[fit.F < 1e-14] = 0 ## they should be 0 (only set to 1e-15 for numerical issue)
```


take a look at the fitted `F`. 
```{r}
hist(log10(fit.F),breaks = 100)
quantile(as.vector(fit.F))
quantile(as.vector(fit.F), probs = seq(0.99, 1, 0.001))

## look at s_k
s_k
```

## simulate data
Look at the oracle $\Lambda$ and data $Y$:
```{r}
p = 1000
k = length(s_k)
idx = sample(x = 1:nrow(fit.F), size = p, replace = FALSE)
Lambda = fit.F[idx,]
Theta = Lambda %*% diag(s_k)
Y = matrix(rpois(n = p * k, lambda = Theta), ncol = k)
hist(log10(Lambda), breaks = 100)
hist(log10(Y), breaks = 100)
```

## Fit with NB-means algorithm (with fast grid search)
```{r cache=TRUE, autodep=TRUE, message=FALSE, warning=FALSE}
verbose = FALSE
maxiter = 1000
fit = nb_means(Y = Y, s = s_k, mu = NULL, A = NULL, maxiter = maxiter, verbose = verbose,
                              control = list(method = "grid"), seed = 123)
```

```{r}
plot(fit$progress)
```


## Compared with MLE result $\lambda^{mle}_{jk} = Y_{jk}/s_k$
```{r}
Lambda_mle = Y %*% diag(1/s_k)
Lambda_eb_pm = fit$posterior$mean ## use posterior mean of the EB result
plot(Lambda, Lambda_eb_pm, xlab = "lambda true", ylab = "lambda fit", col = "blue")
points(Lambda, Lambda_mle, col = "red")
abline(0,1,lwd=0.5,col="black")

legend("bottomright", legend=c("posterior mean (EB)", "MLE"),
       col=c("blue", "red"), lty=c(1,1), cex=0.8)
```
Looks like the dataset is too easy?\

Take a closer look
```{r}
mask <- (Lambda>0)
plot(Lambda[mask], Lambda_eb_pm[mask], log = "xy", xlab = "lambda true", ylab = "lambda fit", col = "blue")
points(Lambda[mask], Lambda_mle[mask], col = "red")
abline(0,1,lwd=0.5,col="black")

legend("topright", legend=c("posterior mean (EB)", "MLE", "y = x"),
       col=c("blue", "red", "black"), lty=c(1,1,1), cex=0.8)
```


Look at some "sparse" rows
```{r}
plot(Lambda[3,], col = "black")
points(Lambda_eb_pm[3,], col = "blue")
points(Lambda_mle[3,], col = "red")
legend("topright", legend=c("truth","posterior mean (EB)", "MLE"),
       col=c("black","blue", "red"), lty=c(1,1), cex=0.8)
```



## look at $\mu$
$\mu$ can be interpreted to be mean for each gene. So I compare it with the true gene mean, and gene median from (true $\Lambda$)
```{r}
gene_mean = apply(Lambda,1, mean)
gene_median = apply(Lambda,1, median)
gene_mean_fitted = fit$fitted_g$mu

plot(gene_mean_fitted, gene_median, col = "red", ylab = "truth mean/median", xlab = "fitted mean")
points(gene_mean_fitted, gene_mean, col = "blue")
abline(0,1,lwd=0.5,col="black")
legend("bottomright", legend=c("gene_median", "gene_mean", "y = x"),
       col=c("red", "blue", "black"), lty=c(1,1,1), cex=0.8)
```


```{r}
hist(gene_median, breaks = 100)
hist(gene_mean, breaks = 100)
hist(gene_mean_fitted, breaks = 100)
```


## look at $\phi_{jk}$
Yusha provided the variability of each gene using Seurat. I look at the Phi of the most and least variable genes (its relaibilty is dubious). There seems not to be much difference. 
```{r}
gene_sub_names = rownames(fit.F)[idx]
gene_ranked = readRDS("./data/genes_ranked.RDS")
id_by_std = order(gene_ranked[gene_sub_names,"variance.standardized"], decreasing = TRUE)
Phi_ordered = 1/fit$fitted_g$A[id_by_std,]

# 
# p1 <- hist(log10(Phi_ordered[1:20,]))                    
# p2 <- hist(log10(Phi_ordered[980:1000,]))                     
# plot( p1, col=rgb(0,0,1,1/4), xlim = c(-5,3), main = "hist of log(phi)")  # first histogram
# plot( p2, col=rgb(1,0,0,1/4), xlim = c(-5,3), add=T)  # second
hist(log10(Phi_ordered[1:20,]), xlim=c(-5,3), col="red", main = "hist of log10(phi)")
hist(log10(Phi_ordered[980:1000,]),add=T, col=rgb(0, 1, 0, 0.5) )
legend("topleft", legend=c("Phi for most variable genes", "Phi for least variable genes"),
       col=c("red", rgb(0, 1, 0, 0.5)), lty=c(1,1), cex=0.8)
```


```{r}
## Phi for most variable genes
table(Phi_ordered[1:20,])
## Phi for least variable genes
table(Phi_ordered[980:1000,])
```








