---
title: "invbeta_gamma"
author: "zihao12"
date: "2020-02-11"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction
The goal is to find a generalization of point-gamma. Naturally we think of beta * gamma:
\begin{align}
& x_i \sim \text{Pois}(\theta_i)\\
& \theta_i = d_i \mu_i\\
& \mu_i \sim \text{Ga}(\alpha, \beta)\\
& d_i \sim \text{Beta}(a, b)
\end{align}



But optimization is hard (need to introduce another variable, and the objective obtained by Variational Inference is not the objective we want in `ebpmf`). 

A conjugate prior is to use GH-Gamma-Poisson. Here I want to study its submodel -- invBeta-gamma-poisson (or PrimeBeta-Gamma-Poisson):

\begin{align}
& x_i \sim \text{Pois}(\theta_i)\\
& \theta_i \sim \text{Ga}(\alpha, \beta_i)\\
& \beta_i \sim \text{invBeta}(a, b)
\end{align}

```{r}
library(extraDistr)
```


## $p(\theta)$ in beta*gamma
```{r}
n = 10000
alpha = 10
beta = 1
a = 0.01
b = 0.01

d = rbeta(n, a, b)
mu = rgamma(n, alpha, rate = beta)
theta = d * mu
plot(density(d))
plot(density(mu))
plot(density(theta))
```
We can get $\theta$ with two modes using `beta * gamma`. Can we do the same un `invBeta-gamma`?

## $p(\theta)$ in invBeta-gamma
To choose appropriate parameters, we want to focus on $\beta_i$. Ideally we $1/\beta_i$ to be of similar shape (two modes) of scaled beta (like its correspondent in beta*gamma)
```{r}
n = 10000
a =20
b = 10
beta = rbetapr(n = n, shape1 = a, shape2 = b)
theta = rgamma(n = n, shape = alpha, rate = beta)
plot(density(1/beta))
plot(density(theta))
```
After playing with a few choices of $(a, b)$, I think it is impossible to get two modes in $p(\beta)$ and $p(\theta)$

## other ways?
We want $1/\beta_i$ to have two modes, where $p(\beta_i)$ has support on $(0,1)$, and we need it be conjugate to gamma. Let's use the shape-scale parameterization for gamma: we want to find $h(.)$ that's two mode, with bounded support,  and is conjugate to Gamma:

\begin{align}
& x_i \sim \text{Pois}(\theta_i)\\
& \theta_i \sim \text{Ga}(\alpha, d_i)\\
& d_i \sim h(.)
\end{align}

```{r}
# beta.gh <- function(psi){
#   psi/(1 - psi)
# }
# 
# psi = rbeta(n = n, shape1 = 0.1, shape2 = 0.1)
# plot(density(psi))
# plot(density(beta.gh(psi)))
```

