---
title: "Issues_ebpmf_rankk"
author: "zihao12"
date: "2019-10-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Experiment setup
It is the same as in https://zihao12.github.io/ebpmf_demo/ebpmf_rankk_demo2.html except that I record the ELBO function. (from `ebpmf::ebpmf_exponential_mixture_experiment` function). 


## Issues:
* ELBO does not strictly increase. I think, if we can really solve each `ebpm` problem optimally within that prior family $\mathcal{G}$, we should mamke ELBO nondecreasing (which is why coordinate ascend works?). But since in practice, we only optimize it over a subset of $\mathcal{G}$ (based on our chosen grid), we may not solve each subproblem well. In fact, the second issue below is a good example. \ 

* Numerical issue when computing `L`: at iteration 107, I get ` Error in verify.likelihood.matrix(L) : 
  Input argument "L" should be a numeric matrix with >= 2 columns, >= 1 rows, all its entries should be non-negative, finite and not NA, and some entries should be positive`. It happens in `ebpm` when computing $L_{ik} = NB(x_i; a_k, \frac{b_k}{b_k + s})$. Only the first column of L (corresponding to the smallest chosen mean (biggest rate)) are all -inf. Why it happens? This is how we choose the grid
```
xprime = x
xprime[x == 0] = xprime[x == 0] + 1
mu_grid_min =  0.05*min(xprime/s)
mu_grid_max = 2*max(x/s)
```
In this case, the smallest nonzero $x$ is very small, so `mu_grid_min` becomes very small ($10^{-18}$). As  a result, the loglikelihood is `-Inf`. So we must revise our way of choosing the grid!!
  

```{r}
library(NNLM)
library(gtools)

sim_mgamma <- function(dist){
  pi = dist$pi
  a = dist$a
  b = dist$b
  idx = which(rmultinom(1,1,pi) == 1)
  return(rgamma(1, shape = a[idx], rate =  b[idx]))
}

## simulate a poisson mean problem
## to do:
simulate_pm  <-  function(n, p, dl, df, K,scale_b = 10, seed = 123){
  set.seed(seed)
  ## simulate L
  a = replicate(dl,1)
  b = 10*runif(dl)
  pi <- rdirichlet(1,rep(1/dl, dl))
  gl = list(pi = pi, a = a, b= b)
  L = matrix(replicate(n*K, sim_mgamma(gl)), ncol = K)
  ## simulate F
  a = replicate(df,1)
  b = 10*runif(df)
  pi <- rdirichlet(1,rep(1/df, df))
  gf = list(pi = pi, a = a, b= b)
  F = matrix(replicate(p*K, sim_mgamma(gf)), ncol = K)
  ## simulate X
  lam = L %*% t(F)
  X = matrix(rpois(n*p, lam), nrow = n)
  Y = matrix(rpois(n*p, lam), nrow = n)
  ## prepare output
  g = list(gl = gl, gf = gf)
  out = list(X = X, Y = Y, L = L, F = F, g = g)
  return(out)
}
```

Simulate data
```{r}
### simulate data
n = 100
p = 200
K = 2
dl = 10
df = 10
scale_b = 5
sim = simulate_pm(n, p, dl, df, K, scale_b = scale_b)
```


Run `ebpmf` and plot ELBO. 
```{r}
## ebpmf
out = ebpmf::ebpmf_exponential_mixture_experiment(sim$X, K, maxiter.out = 100, verbose = F)
elbo_ebpmf= out$ELBO
out_ebpmf = out$qg
#elbo_ebpmf = elbo_ebpmf - min(elbo_ebpmf)
distance_max =  max(elbo_ebpmf) - elbo_ebpmf
plot(1:length(distance_max), distance_max, xlab = "niter", ylab = "distance to max (ELBO)", type = "l", col = "blue")
```









