---
title: "nmf_sparse"
author: "zihao12"
date: "2019-10-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Adapted from https://stephens999.github.io/misc/nmf_sparse.html



```{r warning=F, message=F}
rm(list = ls())
library("NNLM")
library("ebpm")
#devtools::load_all("../ebpm")
library("ebpmf.alpha")
#devtools::load_all("../ebpmf.alpha")
source("code/misc.R")
```



## "easy" data with "block-like structure" 
We simulate data with 3 factors with a "block-like" structure.
```{r}
set.seed(123)
n = 99
p = 300
k= 3
L = matrix(0, nrow=n, ncol=k)
F = matrix(0, nrow=p, ncol=k)
L[1:(n/3),1] = 1
L[((n/3)+1):(2*n/3),2] = 1
L[((2*n/3)+1):n,3] = 1
  
F[1:(p/3),1] = 1+10*runif(p/3)
F[((p/3)+1):(2*p/3),2] = 1+10*runif(p/3)
F[((2*p/3)+1):p,3] = 1+10*runif(p/3)
lambda = L %*% t(F)
X = matrix(rpois(n=length(lambda),lambda),nrow=n)
image(X)
```

Now run MLE and show L, F
```{r warning=F, message=F}
fit_lee = NNLM::nnmf(A = X, k = 3, loss = "mkl", method = "lee", max.iter = 10000)

par(mfrow = c(2,2))
for(i in 1:k){
  plot(fit_lee$W[,i],main=paste0("mle_lee:  loadings ",i), ylab = "loading")
}

par(mfrow = c(2,2))
for(i in 1:k){
  plot(fit_lee$H[i,],main=paste0("mle_lee:  factors ",i), ylab = "factor")
}
```




## Add a dense fourth factor
Now I add a fourth factor that is dense.
Note that we can make the problem much harder
by making the 4th (dense) factor have larger PVE (increase `mfac` in the code).
That may be useful for comparing methods on a harder problem...
```{r cache=TRUE, autodep=TRUE, message=FALSE, warning=FALSE}
set.seed(123)
n = 99
p = 300
k= 4
mfac = 2.5 # controls PVE of dense factor
L = matrix(0, nrow=n, ncol=k)
F = matrix(0, nrow=p, ncol=k)
L[1:(n/3),1] = 1
L[((n/3)+1):(2*n/3),2] = 1
L[((2*n/3)+1):n,3] = 1
L[,4] = 1+mfac*runif(n)
F[1:(p/3),1] = 1+10*runif(p/3)
F[((p/3)+1):(2*p/3),2] = 1+10*runif(p/3)
F[((2*p/3)+1):p,3] = 1+10*runif(p/3)
F[,4]= 1+mfac*runif(p)
lambda = L %*% t(F)
X = matrix(rpois(n=length(lambda),lambda),nrow=n)
image(X)
```



## Fit with  MLE (lee) 
```{r cache=TRUE, autodep=TRUE, message=FALSE, warning=FALSE}
k = 4
fit_lee = NNLM::nnmf(A = X, k = k, loss = "mkl", method = "lee", max.iter = 500, verbose = FALSE)
# qg_mle = ebpmf.alpha::initialize_qg_from_LF(L0 = fit_lee$W, F0 =t( fit_lee$H))
# init_list = list(qg = qg_mle)
```

## Fit with  ebpmf with different prior families 
```{r cache=TRUE, autodep=TRUE, message=FALSE, warning=FALSE}
## run ebpmf methods
### setup
maxiter = 1000
verbose = FALSE
tol = 1e-30

### initialiazation
init_lf = NNLM::nnmf(A = X, k = k, loss = "mkl", method = "scd", max.iter = 20, verbose = FALSE)
init = list(qg = initialize_qg_from_LF(L0 = init_lf$W, F0 = t(init_lf$H)))

### two-gamma (this is very very slow, around 60 tims slower than others)
start = proc.time()
fit_ebpmf_tg = ebpmf.alpha::ebpmf(X, K = k,pm_func = ebpm::ebpm_two_gamma, init = init,
                                  maxiter = maxiter, verbose = verbose, tol = tol)
runtime = proc.time()- start
fit_ebpmf_tg$runtime = runtime[[3]]

### point-gamma
start = proc.time()
fit_ebpmf_pg = ebpmf.alpha::ebpmf(X, K = k,pm_func = ebpm::ebpm_point_gamma, init = init,
                                  maxiter = maxiter, verbose = verbose, tol = tol)
runtime = proc.time()- start
fit_ebpmf_pg$runtime = runtime[[3]]

### invBeta-gamma
start = proc.time()
fit_ebpmf_bg = ebpmf.alpha::ebpmf(X, K = k,pm_func = ebpm::ebpm_invBeta_gamma, init = init,
                                  pm_control = list(fix_g = c(FALSE, FALSE, FALSE)),
                                  maxiter = maxiter, verbose = verbose, tol = tol)
runtime = proc.time()- start
fit_ebpmf_bg$runtime = runtime[[3]]

### gamma-mixture-single-scale
start = proc.time()
fit_ebpmf_gammamix = ebpmf.alpha::ebpmf(X, K = k,pm_func = ebpm::ebpm_gamma_mixture_single_scale, init = init,
                                  maxiter = maxiter, verbose = verbose, tol = tol)
runtime = proc.time()- start
fit_ebpmf_gammamix$runtime = runtime[[3]]

# ### exp-mixture (has a bug, at least when iter = 100)
# start = proc.time()
# fit_ebpmf_expmix = ebpmf.alpha::ebpmf(X, K = k,pm_func = ebpm::ebpm_exponential_mixture, init = init,
#                                   maxiter = maxiter, verbose = verbose, tol = tol)
# # runtime = proc.time()- start
# fit_ebpmf_expmix$runtime = runtime[[3]]


```

<!-- ## compare runtime -->
<!-- ```{r} -->
<!-- t(list(lee = fit_lee$run.time[[3]], tg = fit_ebpmf_tg$runtime, pg = fit_ebpmf_pg$runtime, bg = fit_ebpmf_bg$runtime, gammamix = fit_ebpmf_gammamix$runtime)) -->
<!-- ``` -->

## Look at the loadings:
```{r}
## lee from truth
par(mfrow = c(2,2))
for(i in 1:k){
  plot(L[,i],main=paste0("truth: loadings ",i), ylab = "loading")
}

## lee from truth
par(mfrow = c(2,2))
for(i in 1:k){
  plot(fit_lee$W[,i],main=paste0("mle_lee: loadings ",i), ylab = "loading")
}

## fit_ebpmf_pg
par(mfrow = c(2,2))
for(d in 1:k){
  plot(fit_ebpmf_tg$qg$qls_mean[,d],main=sprintf("ebpmf_two_gamma: loadings %d", d), ylab = "loading")
}


## fit_ebpmf_pg
par(mfrow = c(2,2))
for(d in 1:k){
  plot(fit_ebpmf_pg$qg$qls_mean[,d],main=sprintf("ebpmf_point_gamma: loadings %d", d), ylab = "loading")
}

## fit_ebpmf_bg
par(mfrow = c(2,2))
for(d in 1:k){
  plot(fit_ebpmf_bg$qg$qls_mean[,d],main=sprintf("ebpmf_invBeta_gamma: loadings %d", d), ylab = "loading")
}

## fit_ebpmf_gammamix
par(mfrow = c(2,2))
for(d in 1:k){
  plot(fit_ebpmf_gammamix$qg$qls_mean[,d],main=sprintf("ebpmf_gammamix(single_scale): loadings %d", d), ylab = "loading")
}
```


## Look at the factors:
```{r}
## lee from truth
par(mfrow = c(2,2))
for(i in 1:k){
  plot(F[,i],main=paste0("truth: factors ",i), ylab = "factors")
}

## lee from truth
par(mfrow = c(2,2))
for(i in 1:k){
  plot(t(fit_lee$H)[,i],main=paste0("mle_lee: factors ",i), ylab = "factors")
}

## fit_ebpmf_pg
par(mfrow = c(2,2))
for(d in 1:k){
  plot(fit_ebpmf_tg$qg$qfs_mean[,d],main=sprintf("ebpmf_two_gamma: factors %d", d), ylab = "factors")
}


## fit_ebpmf_pg
par(mfrow = c(2,2))
for(d in 1:k){
  plot(fit_ebpmf_pg$qg$qfs_mean[,d],main=sprintf("ebpmf_point_gamma: factors %d", d), ylab = "factors")
}

## fit_ebpmf_bg
par(mfrow = c(2,2))
for(d in 1:k){
  plot(fit_ebpmf_bg$qg$qfs_mean[,d],main=sprintf("ebpmf_invBeta_gamma: factors %d", d), ylab = "factors")
}

## fit_ebpmf_gammamix
par(mfrow = c(2,2))
for(d in 1:k){
  plot(fit_ebpmf_gammamix$qg$qfs_mean[,d],main=sprintf("ebpmf_gammamix(single_scale): factors %d", d), ylab = "factors")
}
```





## compare divergence from truth
```{r}
source("code/misc.R")
lam_mle = fit_lee$W %*% fit_lee$H
lam_ebpmf_tg = fit_ebpmf_tg$qg$qls_mean %*% t(fit_ebpmf_tg$qg$qfs_mean)
lam_ebpmf_pg = fit_ebpmf_pg$qg$qls_mean %*% t(fit_ebpmf_pg$qg$qfs_mean)
lam_ebpmf_bg = fit_ebpmf_bg$qg$qls_mean %*% t(fit_ebpmf_bg$qg$qfs_mean)
lam_ebpmf_gammamix = fit_ebpmf_gammamix$qg$qls_mean %*% t(fit_ebpmf_gammamix$qg$qfs_mean)

rmse = c(0, RMSE(lambda, lam_mle), RMSE(lambda, lam_ebpmf_tg), RMSE(lambda, lam_ebpmf_pg),
         RMSE(lambda, lam_ebpmf_bg), RMSE(lambda, lam_ebpmf_gammamix))
js = c(0, JS(lambda, lam_mle), JS(lambda, lam_ebpmf_tg), JS(lambda, lam_ebpmf_pg),
         JS(lambda, lam_ebpmf_bg), JS(lambda, lam_ebpmf_gammamix))
kl = c(0, KL(lambda, lam_mle), KL(lambda, lam_ebpmf_tg), KL(lambda, lam_ebpmf_pg),
         KL(lambda, lam_ebpmf_bg), KL(lambda, lam_ebpmf_gammamix))
ll = c(sum(dpois(X, L %*% t(F),log=TRUE)),
       sum(dpois(X, fit_lee$W %*% fit_lee$H,log=TRUE)),
       sum(dpois(X, fit_ebpmf_tg$qg$qls_mean %*% t(fit_ebpmf_tg$qg$qfs_mean),log=TRUE)),
       sum(dpois(X, fit_ebpmf_pg$qg$qls_mean %*% t(fit_ebpmf_pg$qg$qfs_mean),log=TRUE)),
       sum(dpois(X, fit_ebpmf_bg$qg$qls_mean %*% t(fit_ebpmf_bg$qg$qfs_mean),log=TRUE)),
       sum(dpois(X, fit_ebpmf_gammamix$qg$qls_mean %*% t(fit_ebpmf_gammamix$qg$qfs_mean),log=TRUE)))

elbo = c(NA, NA, fit_ebpmf_tg$ELBO[length(fit_ebpmf_tg$ELBO)],  fit_ebpmf_pg$ELBO[length(fit_ebpmf_pg$ELBO)],
          fit_ebpmf_bg$ELBO[length(fit_ebpmf_bg$ELBO)],  fit_ebpmf_gammamix$ELBO[length(fit_ebpmf_gammamix$ELBO)])

runtime = c(NA, fit_lee$run.time[[3]], fit_ebpmf_tg$runtime, fit_ebpmf_pg$runtime, 
            fit_ebpmf_bg$runtime, fit_ebpmf_gammamix$runtime)

dist = data.frame(cbind(rmse, js, kl, ll, elbo, runtime), row.names = c("oracle","mle", "two-gamma" ,"point-gamma", "invBeta-gamma", "gammamix"))
dist
```

