---
title: "nmf_sparse6"
author: "zihao12"
date: "2019-11-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
Dongyue did an investigation into some penalty-based methods in https://dongyuexie.github.io/SMF/nmfsparse.html. I copy them here and look at their loglikelihood and see how well they fit the data. 

## Summary
* Those methods induce sparsity but have very low loglikelihood (compared to truth). The  reason is that there are too many fits with $x >> \hat{\lambda}$ (not many exact 0s though). Also, when making penalty parameter large, the algorithm gives warning  that "beta is too large", probably because of the bad behavior of Poisson loglikelihood when  $\lambda$ is near 0 . \

* With this observation, it makes sense  to introduce random effect to  this model: $X_{ij} = Pois(\sum_k l_{ik} f_{jk} + \mu_{ij})$, and we hope the model  can set some $L, F$ to 0, and let $\mu_{ij}$ take care of the nonzero-small fit.\

* We initialize `ebpmf_random_effect` with the results from `nsNMF`, with $\mu$ initialized to those underfits part $(X_{ij} - (LF^t)_{ij})_{+}$(not sure if it makes sense). In the experiment in the end we get shrinkage effect as point gamma works as expected. 

## Data

```{r}
log_lik <- function(X, lam){
  return(sum(dpois(x = X, lambda = lam , log = T)))
}
```


```{r}
set.seed(123)
n = 99
p = 300
k= 4
mfac = 2 # controls PVE of dense factor
L = matrix(0, nrow=n, ncol=k)
F = matrix(0, nrow=p, ncol=k)

L[1:(n/3),1] = 1
L[((n/3)+1):(2*n/3),2] = 1
L[((2*n/3)+1):n,3] = 1
L[,4] = 1+mfac*runif(n)

F[1:(p/3),1] = 1+10*runif(p/3)
F[((p/3)+1):(2*p/3),2] = 1+10*runif(p/3)
F[((2*p/3)+1):p,3] = 1+10*runif(p/3)
F[,4]= 1+mfac*runif(p)

lambda = L %*% t(F)
X = matrix(rpois(n=length(lambda),lambda),nrow=n)

log_lik(X, lam = lambda)
image(X)

```


### Non-smooth NMF

Non-smooth NMF. Uses a modified version of Lee and Seung’s multiplicative updates for Kullback-Leibler divergence to fit a extension of the standard NMF model. It is meant to give sparser results.

Reference: [Pascual-Montano2006](https://ieeexplore.ieee.org/document/1580485)

(Note the column sum of `W` are all 1s. Think we prefer the column sum of `F` to be constant?)

```{r warning=F, message=F}

library(NMF)

fit_nsNMF = nmf(X,4,method = 'nsNMF')
log_lik(X, lam = fit_nsNMF@fit@W %*% fit_nsNMF@fit@H)

par(mfrow=c(2,2))
for(i in 1:k){
    plot(fit_nsNMF@fit@W[,i],main=paste0("nsNMF: estimated loadings ",i))
}

par(mfrow=c(2,2))
for(i in 1:k){
    plot(fit_nsNMF@fit@H[i,],main=paste0("nsNMF: estimated factors ",i))
}


```

## Sparse NMF

Alternating Least Square (ALS) approach. It is meant to be very fast com- pared to other approaches.

Reference: [KimH2007](https://www.ncbi.nlm.nih.gov/pubmed/17483501)

Sparse Loading:

```{r}

fit_snmfl = nmf(X,4,method = 'snmf/l',beta=20)

par(mfrow=c(2,2))
for(i in 1:k){
    plot(fit_snmfl@fit@W[,i],main=paste0("snmfl: estimated loadings ",i))
}

par(mfrow=c(2,2))
for(i in 1:k){
    plot(fit_snmfl@fit@H[i,],main=paste0("snmfl: estimated factors ",i))
}


log_lik(X, lam = fit_snmfl@fit@W %*% fit_snmfl@fit@H)

```


Sparse Factors:

```{r}
fit_snmfr = nmf(X,4,method = 'snmf/r',beta=1)

par(mfrow=c(2,2))
for(i in 1:k){
    plot(fit_snmfr@fit@W[,i],main=paste0("snmfr: estimated loadings ",i))
}

par(mfrow=c(2,2))
for(i in 1:k){
    plot(fit_snmfr@fit@H[i,],main=paste0("snmfr: estimated factors ",i))
}

log_lik(X, lam = fit_snmfr@fit@W %*% fit_snmfr@fit@H)

```

The loglikelihood from those models are often very bad. 

## `ebpmf` (initialized from truth)
I think  it is a good fit, with sparsity as well as good loglikelihood and divergence from true lambda. We would believe a good sparse model should get a simialr likelihood.
```{r warning=F, echo =  F}
devtools::load_all("../ebpmf.alpha")
devtools::load_all("../ebpm")
library(ebpmf)
qg_from_truth = initialize_qg_from_LF(L = L, F = F)
# fit_pg_from_truth = ebpmf::ebpmf_point_gamma(X = X, K = k, qg = qg_from_truth, maxiter.out = 100)
# saveRDS(fit_pg_from_truth, "data/nmf_sparse6_fit_pg_from_truth.Rds")
fit_pg_from_truth = readRDS("data/nmf_sparse6_fit_pg_from_truth.Rds")
log_lik(X, lam = fit_pg_from_truth$qg$qls_mean %*% t(fit_pg_from_truth$qg$qfs_mean))
```


## See why those sparse-NMF methods have low loglikelihood
```{r}
lam_nsnmf = fit_nsNMF@fit@W %*% fit_nsNMF@fit@H
lam_ebpm = fit_pg_from_truth$qg$qls_mean %*% t(fit_pg_from_truth$qg$qfs_mean)

par(mfrow=c(2,2))
plot(X, lam_nsnmf)
plot(X, lambda)
plot(lambda, lam_nsnmf)
```

Clearly, there are many counts with $x >> \hat{\lambda}$. This may help us  initialize `ebpmf_random_effect`


## Fit with  `ebpmf_random_effect`
* The model is $X_{ij} = Pois(\sum_k l_{ik} f_{jk} + \mu_{ij})$, with prior on each column of $L, F$, and on entire $\mu_{ij}$. It is fitted very similarly to `ebpmf`, except that we introduce $Z^0$ with $Z^0_{ij} \sim Pois(\mu_{ij})$. Now I use point gamma for the posterior family for all of them. \

* Initialization: I need to initialize $q$ for $L, F, \mu$. I first fit the model with `nsNMF`, then with  the  `L`, `F` (after thresholding some small ones to be 0, as otherwise $\pi_0$ will be all 0 again) I set $\mu_{ij} =  (X_{ij} - (LF^t)_{ij})_{+}$. Then I use the  point  estimate as  posterior mean, and log  of it to be  posterior log-mean as initialization. 

```{r}
initialize_qg_from_LF_random_effect <- function(L0,F0, mu){
  K = ncol(L0)
  qls_mean = L0
  qfs_mean = F0
  qls_mean_log = log(L0)
  qfs_mean_log = log(F0)
  qmu_mean = mu
  qmu_mean_log = log(mu)
  qg = list(qls_mean = qls_mean, qls_mean_log =qls_mean_log,
            qfs_mean = qfs_mean, qfs_mean_log =qfs_mean_log,
            qmu_mean = qmu_mean, qmu_mean_log = qmu_mean_log,
            gls = replicate(K, list(NULL)),gfs = replicate(K, list(NULL)))
  return(qg)
}
 
L0 = fit_nsNMF@fit@W
F0 = t(fit_nsNMF@fit@H)

thres <- function(x, cut = 0.1){
  x[x < cut*mean(x)] = 0
  return(x)
}

L0_thres = L0
L0_thres = apply(L0_thres, 2, FUN = thres)

F0_thres = F0
F0_thres = apply(F0_thres, 2, FUN = thres)

par(mfrow=c(1,2))
plot(L0, L0_thres)
plot(F0, F0_thres)

mu = X - L0_thres %*%  t(F0_thres)
mu[mu < 0] = 0

print(sum(mu ==  0)/(n*p))

qg_from_nsNMF = initialize_qg_from_LF_random_effect(L0_thres, F0_thres, mu)
```


```{r}
# fit_ebpmf_re_nsnmf = ebpmf_random_effect(X = X, K = k, maxiter.out = 100, qg = qg_from_nsNMF)
# saveRDS(fit_ebpmf_re_nsnmf, "data/nmf_sparse6_fit_ebpmf_re_nsnmf.Rds")
fit_ebpmf_re_nsnmf = readRDS("data/nmf_sparse6_fit_ebpmf_re_nsnmf.Rds")
```


```{r}
lam_ebpmf_re = fit_ebpmf_re_nsnmf$qg$qls_mean %*% t(fit_ebpmf_re_nsnmf$qg$qfs_mean) + fit_ebpmf_re_nsnmf$qg$qmu_mean
log_lik(X, lam_ebpmf_re)

for(idx in 1:k){
  par(mfrow=c(1,2))
  plot(L0_thres[,idx],fit_ebpmf_re_nsnmf$qg$qls_mean[,idx], 
       xlab = sprintf("L0_thres: loading %d", idx), ylab = sprintf("L_ebpmf:loading %d", idx), col  = "red")
  abline(a = 0, b = 1)
  plot(F0_thres[,idx],fit_ebpmf_re_nsnmf$qg$qfs_mean[,idx], 
       xlab = sprintf("F0_thres: factor %d", idx), ylab = sprintf("F_ebpmf:factor %d", idx), col  = "blue")
  abline(a = 0, b = 1)
}
```
Finally we have some shrinkage effects!\

Take a look at the corresponding $Z$. ($EZ[,,5]$ corresponds to $\mu$).  
```{r}
Ez_ebpmf_random_effect = get_Ez_random_effect(X = X, qg = fit_ebpmf_re_nsnmf$qg, K = k)$Ez
for(i in 1:(k+1)){
 image(Ez_ebpmf_random_effect[,,i], main = sprintf("Ez[,,%d]", i)) 
}
```


