---
title: "nmf_sparse"
author: "zihao12"
date: "2019-10-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

Adapted from https://stephens999.github.io/misc/nmf_sparse.html



```{r warning=F, message=F}
rm(list = ls())
library("NNLM")
library("ebpm")
library("ebpmf")
source("code/misc.R")
```

## Introduction
The goal is to do some simple simulations where the factors are sparse
and look at the sparsity of the solutions from regular (unpenalized) nmf.

## Result:
* After we add a dense fourth factor, we can see `ebpmf_two_gamma` obtains "better" loadings. \
* current `ebpmf_two_gamma` is much slower to fit

We simulate data with 3 factors with a "block-like"" structure.
```{r}
set.seed(123)
n = 99
p = 300
k= 3
L = matrix(0, nrow=n, ncol=k)
F = matrix(0, nrow=p, ncol=k)
L[1:(n/3),1] = 1
L[((n/3)+1):(2*n/3),2] = 1
L[((2*n/3)+1):n,3] = 1
  
F[1:(p/3),1] = 1+10*runif(p/3)
F[((p/3)+1):(2*p/3),2] = 1+10*runif(p/3)
F[((2*p/3)+1):p,3] = 1+10*runif(p/3)
lambda = L %*% t(F)
X = matrix(rpois(n=length(lambda),lambda),nrow=n)
image(X)
```

Now run MLE
```{r warning=F, message=F}
fit_lee = NNLM::nnmf(A = X, k = 3, loss = "mkl", method = "lee", max.iter = 10000)

for(i in 1:k){
  plot(fit_lee$W[,i],main=paste0("mle_lee: estimated loadings ",i))
}
```




## Add a dense fourth factor

Now I add a fourth factor that is dense.
Note that we can make the problem much harder
by making the 4th (dense) factor have larger PVE (increase `mfac` in the code).
That may be useful for comparing methods on a harder problem...
```{r cache=TRUE, autodep=TRUE, message=FALSE, warning=FALSE}
set.seed(123)
n = 99
p = 300
k= 4
mfac = 2.5 # controls PVE of dense factor
L = matrix(0, nrow=n, ncol=k)
F = matrix(0, nrow=p, ncol=k)
L[1:(n/3),1] = 1
L[((n/3)+1):(2*n/3),2] = 1
L[((2*n/3)+1):n,3] = 1
L[,4] = 1+mfac*runif(n)
F[1:(p/3),1] = 1+10*runif(p/3)
F[((p/3)+1):(2*p/3),2] = 1+10*runif(p/3)
F[((2*p/3)+1):p,3] = 1+10*runif(p/3)
F[,4]= 1+mfac*runif(p)
lambda = L %*% t(F)
X = matrix(rpois(n=length(lambda),lambda),nrow=n)
image(X)
```



### Fit with  MLE (lee) , ebpmf (point-gamma & two-gamma). 
```{r cache=TRUE, autodep=TRUE, message=FALSE, warning=FALSE}
k = 4
fit_lee = NNLM::nnmf(A = X, k = k, loss = "mkl", method = "lee", max.iter = 10000, verbose = FALSE)
fit_ebpmf_pg = ebpmf.alpha::ebpmf_point_gamma(X, K = k, maxiter.out = 100, verbose = FALSE)
fit_ebpmf_tg = ebpmf.alpha::ebpmf_two_gamma(X, K = k, maxiter.out = 100, verbose = FALSE)

## ll (oracle)
sum(dpois(X, L %*% t(F),log=TRUE))
## ll (mle-lee)
sum(dpois(X, fit_lee$W %*% fit_lee$H,log=TRUE))
## ll (ebpmf-point-gamma)
sum(dpois(X, fit_ebpmf_pg$qg$qls_mean %*% t(fit_ebpmf_pg$qg$qfs_mean),log=TRUE))
## ll (ebpmf-two-gamma)
sum(dpois(X, fit_ebpmf_tg$qg$qls_mean %*% t(fit_ebpmf_tg$qg$qfs_mean),log=TRUE))
```


### Look at the loadings:
```{r}
## lee from truth
par(mfrow = c(2,2))
for(i in 1:k){
  plot(fit_lee$W[,i],main=paste0("mle_lee: loadings ",i), ylab = "loading")
}
## ebpmf_pg 
for(d in 1:k){
  plot(fit_ebpmf_pg$qg$qls_mean[,d],main=sprintf("ebpmf_point_gamma: loadings %d", d), ylab = "loading")
}
## ebpmf_tg 
for(d in 1:k){
  plot(fit_ebpmf_tg$qg$qls_mean[,d],main=sprintf("ebpmf_two_gamma: loadings %d", d), ylab = "loading")
}
```

### Look at the factors:
```{r}
## lee from truth
par(mfrow = c(2,2))
for(i in 1:k){
  plot(fit_lee$H[i,],main=paste0("mle_lee: factors ",i), ylab = "factor")
}
## ebpmf_pg 
for(d in 1:k){
  plot(fit_ebpmf_pg$qg$qfs_mean[,d],main=sprintf("ebpmf_point_gamma: factors %d", d), ylab = "factor")
}
## ebpmf_tg 
for(d in 1:k){
  plot(fit_ebpmf_tg$qg$qfs_mean[,d],main=sprintf("ebpmf_two_gamma: factors %d", d), ylab = "factor")
}
```



Compare $\Lambda_{est}$ with $\Lambda_{true}$
```{r}
lam_mle = fit_lee$W %*% fit_lee$H
lam_ebpmf_pg = fit_ebpmf_pg$qg$qls_mean %*% t(fit_ebpmf_pg$qg$qfs_mean)
lam_ebpmf_tg = fit_ebpmf_tg$qg$qls_mean %*% t(fit_ebpmf_tg$qg$qfs_mean)

rmse = c(RMSE(lambda, lam_mle), RMSE(lambda, lam_ebpmf_pg), RMSE(lambda, lam_ebpmf_tg))
js = c(JS(lambda, lam_mle), JS(lambda, lam_ebpmf_pg), JS(lambda, lam_ebpmf_tg))
kl = c(KL(lambda, lam_mle), KL(lambda, lam_ebpmf_pg), KL(lambda, lam_ebpmf_tg))

dist = data.frame(cbind(rmse, js, kl), row.names = c("mle", "point-gamma", "two-gamma"))
dist
```
























