---
title: "nmf_sparse"
author: "zihao12"
date: "2019-10-30"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Goal and Description
* Previous results show that `ebpmf-two-gamma` can recover the sparse truth very well. However it is too slow. \

* Original `ebpm-two-gamma` runs EM algorithm for the MLE part, and terminates when parameters change little.\

* I speed up each EM iteration (see [comparison](https://zihao12.github.io/ebpmf_demo/compare_ebpm_two_gamma_speed.html#result_summary)), and restrict the number of EM iterations. \

## Result
* [Runtime](https://zihao12.github.io/ebpmf_demo/nmf_sparse16.html#compare_runtime): `ebpmf-two-gamma_fast` is comparable to `ebpmf-point-gamma`\

* Fit: however fit is worse. One sparse loading is [not recovered](https://zihao12.github.io/ebpmf_demo/nmf_sparse16.html#look_at_the_loadings:), and the recoevered mean is further from truth compared to `ebpmf_two_gamma_slow`. \

* Look at the [ELBO](https://zihao12.github.io/ebpmf_demo/nmf_sparse16.html#compare_elbo), we can see the restricted number of iterations prevent the algorithm from getting a good solution. \

* Will try different stopping criteria for each `ebpm_two_gamma` problem. 

```{r warning=F, message=F}
rm(list = ls())
library("NNLM")
library("ebpm")
#devtools::load_all("../ebpm")
library("ebpmf.alpha")
#devtools::load_all("../ebpmf.alpha")
source("code/misc.R")
```

## Simulate data
```{r cache=TRUE, autodep=TRUE, message=FALSE, warning=FALSE}
set.seed(123)
n = 99
p = 300
k= 4
mfac = 2.5 # controls PVE of dense factor
L = matrix(0, nrow=n, ncol=k)
F = matrix(0, nrow=p, ncol=k)
L[1:(n/3),1] = 1
L[((n/3)+1):(2*n/3),2] = 1
L[((2*n/3)+1):n,3] = 1
L[,4] = 1+mfac*runif(n)
F[1:(p/3),1] = 1+10*runif(p/3)
F[((p/3)+1):(2*p/3),2] = 1+10*runif(p/3)
F[((2*p/3)+1):p,3] = 1+10*runif(p/3)
F[,4]= 1+mfac*runif(p)
lambda = L %*% t(F)
X = matrix(rpois(n=length(lambda),lambda),nrow=n)
```

## Fit with  MLE and ebpmf methods
```{r cache=TRUE, autodep=TRUE, message=FALSE, warning=FALSE}
## run ebpmf methods
### setup
maxiter = 1000
verbose = FALSE
tol = -1

## Fit with  MLE (lee) 
k = 4
fit_lee = NNLM::nnmf(A = X, k = k, loss = "mkl", method = "lee", max.iter = maxiter, verbose = FALSE)

### initialiazation
init_lf = NNLM::nnmf(A = X, k = k, loss = "mkl", method = "scd", max.iter = 20, verbose = FALSE)
init = list(qg = initialize_qg_from_LF(L0 = init_lf$W, F0 = t(init_lf$H)))

### point-gamma
start = proc.time()
fit_ebpmf_pg = ebpmf.alpha::ebpmf(X, K = k,pm_func = ebpm::ebpm_point_gamma, init = init,
                                  maxiter = maxiter, verbose = verbose, tol = tol)
runtime = proc.time()- start
fit_ebpmf_pg$runtime = runtime[[3]]


### two-gamma (faster version)
start = proc.time()
fit_ebpmf_tg = ebpmf.alpha::ebpmf(X, K = k,pm_func = ebpm::ebpm_two_gamma_fast3, 
                                  pm_control = list(n_iter = 10, control = list(tol_in = 1e-2)),
                                  init = init, maxiter = maxiter, verbose = verbose, tol = tol)
runtime = proc.time()- start
fit_ebpmf_tg$runtime = runtime[[3]]

### two-gamma (this is very very slow, around 60 tims slower than others)
start = proc.time()
fit_ebpmf_tg_slow = ebpmf.alpha::ebpmf(X, K = k,pm_func = ebpm::ebpm_two_gamma, init = init,
                                  maxiter = maxiter, verbose = verbose, tol = tol)
runtime = proc.time()- start
fit_ebpmf_tg_slow$runtime = runtime[[3]]
```

## compare runtime
```{r}
t(list(lee = fit_lee$run.time[[3]], tg_slow = fit_ebpmf_tg_slow$runtime, tg = fit_ebpmf_tg$runtime, pg = fit_ebpmf_pg$runtime))
```

## Look at the loadings:
```{r}
## lee from truth
par(mfrow = c(2,2))
for(i in 1:k){
  plot(L[,i],main=paste0("truth: loadings ",i), ylab = "loading")
}

## lee from truth
par(mfrow = c(2,2))
for(i in 1:k){
  plot(fit_lee$W[,i],main=paste0("mle_lee: loadings ",i), ylab = "loading")
}

## fit_ebpmf_tg_slow
par(mfrow = c(2,2))
for(d in 1:k){
  plot(fit_ebpmf_tg_slow$qg$qls_mean[,d],main=sprintf("ebpmf_two_gamma (slow): loadings %d", d), ylab = "loading")
}


## fit_ebpmf_tg
par(mfrow = c(2,2))
for(d in 1:k){
  plot(fit_ebpmf_tg$qg$qls_mean[,d],main=sprintf("ebpmf_two_gamma (fast): loadings %d", d), ylab = "loading")
}


## fit_ebpmf_pg
par(mfrow = c(2,2))
for(d in 1:k){
  plot(fit_ebpmf_pg$qg$qls_mean[,d],main=sprintf("ebpmf_point_gamma: loadings %d", d), ylab = "loading")
}
```


## compare divergence from truth
```{r}
source("code/misc.R")
lam_mle = fit_lee$W %*% fit_lee$H
lam_ebpmf_tg_slow = fit_ebpmf_tg_slow$qg$qls_mean %*% t(fit_ebpmf_tg_slow$qg$qfs_mean)
lam_ebpmf_tg = fit_ebpmf_tg$qg$qls_mean %*% t(fit_ebpmf_tg$qg$qfs_mean)
lam_ebpmf_pg = fit_ebpmf_pg$qg$qls_mean %*% t(fit_ebpmf_pg$qg$qfs_mean)

rmse = c(0, RMSE(lambda, lam_mle), RMSE(lambda, lam_ebpmf_tg_slow),RMSE(lambda, lam_ebpmf_tg), RMSE(lambda, lam_ebpmf_pg))
js = c(0, JS(lambda, lam_mle), JS(lambda, lam_ebpmf_tg_slow),JS(lambda, lam_ebpmf_tg), JS(lambda, lam_ebpmf_pg))
kl = c(0, KL(lambda, lam_mle), KL(lambda, lam_ebpmf_tg_slow), KL(lambda, lam_ebpmf_tg), KL(lambda, lam_ebpmf_pg))
ll = c(sum(dpois(X, L %*% t(F),log=TRUE)),
       sum(dpois(X, fit_lee$W %*% fit_lee$H,log=TRUE)),
       sum(dpois(X, fit_ebpmf_tg_slow$qg$qls_mean %*% t(fit_ebpmf_tg_slow$qg$qfs_mean),log=TRUE)),
       sum(dpois(X, fit_ebpmf_tg$qg$qls_mean %*% t(fit_ebpmf_tg$qg$qfs_mean),log=TRUE)),
       sum(dpois(X, fit_ebpmf_pg$qg$qls_mean %*% t(fit_ebpmf_pg$qg$qfs_mean),log=TRUE)))


elbo = c(NA, NA, fit_ebpmf_tg_slow$ELBO[length(fit_ebpmf_tg_slow$ELBO)],fit_ebpmf_tg$ELBO[length(fit_ebpmf_tg$ELBO)],  fit_ebpmf_pg$ELBO[length(fit_ebpmf_pg$ELBO)])

data.frame(cbind(rmse, js, kl, ll, elbo), row.names = c("oracle","mle", "two-gamma_slow","two-gamma" ,"point-gamma"))
```

## Compare ELBO
```{r}
plot(fit_ebpmf_tg_slow$ELBO, ,col = "red", xlab = "niter", ylab = "ELBO")
lines(fit_ebpmf_tg$ELBO, col = "blue")
legend("bottomright", legend=c("slow", "fast"),
       col=c("red", "blue"), lty=2:1, cex=0.8)
```

