---
title: "nmf_sparse6"
author: "zihao12"
date: "2019-11-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
From https://dongyuexie.github.io/SMF/nmfsparse.html

## Summary

1. L1 penalty in NNLM::nnmf either put no/littlle thresholding or thresholds all entries to zero.

2. nsNMF and snmf in NMF package give sparse estimate of loadings and factors.

## Data

```{r}
log_lik <- function(X, lam){
  return(sum(dpois(x = X, lambda = lam , log = T)))
}
```


```{r}
set.seed(123)
n = 99
p = 300
k= 4
mfac = 2 # controls PVE of dense factor
L = matrix(0, nrow=n, ncol=k)
F = matrix(0, nrow=p, ncol=k)

L[1:(n/3),1] = 1
L[((n/3)+1):(2*n/3),2] = 1
L[((2*n/3)+1):n,3] = 1
L[,4] = 1+mfac*runif(n)

F[1:(p/3),1] = 1+10*runif(p/3)
F[((p/3)+1):(2*p/3),2] = 1+10*runif(p/3)
F[((2*p/3)+1):p,3] = 1+10*runif(p/3)
F[,4]= 1+mfac*runif(p)

lambda = L %*% t(F)
X = matrix(rpois(n=length(lambda),lambda),nrow=n)

log_lik(X, lam = lambda)

image(X)

```

## NNLM-nnmf

```{r,warning=FALSE}
library(NNLM)

fit_scd = NNLM::nnmf(A = X, k = 4, loss = "mse", method = "scd", max.iter = 10000)

log_lik(X, lam = fit_scd$W %*% fit_scd$H)



fit_scd_L1.1 = NNLM::nnmf(A = X, k = 4, loss = "mse", method = "scd", max.iter = 10000, alpha=c(0,0,1))
log_lik(X, lam = fit_scd_L1.1$W %*% fit_scd_L1.1$H)

par(mfrow=c(2,2))
for(i in 1:k){
    plot(fit_scd_L1.1$W[,i],main=paste0("L1 penalty = 1: estimated loadings ",i))
}


fit_scd_L1.10 = NNLM::nnmf(A = X, k = 4, loss = "mse", method = "scd", max.iter = 10000, alpha=c(0,0,10))
log_lik(X, lam = fit_scd_L1.10$W %*% fit_scd_L1.10$H)

par(mfrow=c(2,2))
for(i in 1:k){
    plot(fit_scd_L1.10$W[,i],main=paste0("L1 penalty = 10: estimated loadings ",i))
}

fit_scd_L1.11 = NNLM::nnmf(A = X, k = 4, loss = "mse", method = "scd", max.iter = 10000, alpha=c(0,0,11))
log_lik(X, lam = fit_scd_L1.11$W %*% fit_scd_L1.11$H)


par(mfrow=c(2,2))
for(i in 1:k){
    plot(fit_scd_L1.11$W[,i],main=paste0("L1 penalty = 11: estimated loadings ",i))
}

fit_scd_L1.100 = NNLM::nnmf(A = X, k = 4, loss = "mse", method = "scd", max.iter = 10000, alpha=c(0,0,100))
log_lik(X, lam = fit_scd_L1.100$W %*% fit_scd_L1.100$H)

par(mfrow=c(2,2))
for(i in 1:k){
    plot(fit_scd_L1.100$W[,i],main=paste0("L1 penalty = 100: estimated loadings ",i))
}
```

## NMF-nmf

### Non-smooth NMF

Non-smooth NMF. Uses a modified version of Lee and Seung’s multiplicative updates for Kullback-Leibler divergence to fit a extension of the standard NMF model. It is meant to give sparser results.

Reference: [Pascual-Montano2006](https://ieeexplore.ieee.org/document/1580485)

```{r}

library(NMF)

fit_nsNMF = nmf(X,4,method = 'nsNMF')
log_lik(X, lam = fit_nsNMF@fit@W %*% fit_nsNMF@fit@H)

par(mfrow=c(2,2))
for(i in 1:k){
    plot(fit_nsNMF@fit@W[,i],main=paste0("nsNMF: estimated loadings ",i))
}

par(mfrow=c(2,2))
for(i in 1:k){
    plot(fit_nsNMF@fit@H[i,],main=paste0("nsNMF: estimated factors ",i))
}


```

## Sparse NMF

Alternating Least Square (ALS) approach. It is meant to be very fast com- pared to other approaches.

Reference: [KimH2007](https://www.ncbi.nlm.nih.gov/pubmed/17483501)

Sparse Loading:

```{r}

fit_snmfl = nmf(X,4,method = 'snmf/l',beta=1)

par(mfrow=c(2,2))
for(i in 1:k){
    plot(fit_snmfl@fit@W[,i],main=paste0("snmfl: estimated loadings ",i))
}

log_lik(X, lam = fit_snmfl@fit@W %*% fit_snmfl@fit@H)

```


Sparse Factors:

```{r}
fit_snmfr = nmf(X,4,method = 'snmf/r',beta=1)

par(mfrow=c(2,2))
for(i in 1:k){
    plot(fit_snmfr@fit@H[i,],main=paste0("snmfr: estimated factors ",i))
}




log_lik(X, lam = fit_snmfr@fit@W %*% fit_snmfr@fit@H)

```



## `ebpmf` (from truth)
```{r warning=F, echo =  F}
devtools::load_all("../ebpmf.alpha")
devtools::load_all("../ebpm")
library(ebpmf)
qg_from_truth = initialize_qg_from_LF(L = L, F = F)
fit_pg_from_truth = ebpmf::ebpmf_point_gamma(X = X, K = k, qg = qg_from_truth, maxiter.out = 100)
log_lik(X, lam = fit_pg_from_truth$qg$qls_mean %*% t(fit_pg_from_truth$qg$qfs_mean))
```







