---
title: "pois_mode_est"
author: "Zihao Wang"
date: "2019-10-04"
header-includes:
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Description  and summary
* I copied from https://stephens999.github.io/misc/pois_mode_est.html. I fixed the bug  that Matthew found and continued comparison. (`ashr` seems not to be updated in github so the result here isn't as good as Matthew's result)

* The data is from https://users.rcc.uchicago.edu/~aksarkar/singlecell-modes/poisson.html#orga229ff0

Basically, the model is

$$
\begin{align}
  & x_i \sim Pois(s_i \lambda_i)\\
  & \lambda_i \sim \delta_{\mu}(.)\\
\end{align}
$$
Then  we have $\hat{\mu} = \frac{\sum_i x_i}{\sum_i s_i}$. So we expect to see our fitted prior to be close to a point mass close to $\hat{\mu}$. 


Some results:\

* loglikelihood: `ebpm_point_gamma` (-2008) > `ashr_pois` (-2097) > `ebpm_exponential_mixture` (-2350)\

* Although `ebpm_exponential_mixture` puts all weight on one exponential distribution, whose mean is close to $\hat{\mu}$, it does not look like a point mass due to the shape constraint of exponential (variance is not small enough).  


```{r}
library(ashr)
library(ebpm)
```

## Data summary

```{r}
d = readRDS("data/pois-mode-est.Rds")
print(sprintf("muhat: %5e", sum(d$x)/sum(d$s)))
hist(d$x)
summary(d$s)
```

## `ashr_pois`
```{r}
res.ash = ash_pois(d$x,d$s,link="identity")
```


```{r echo = F}
print(sprintf("loglikelihood:  %f", res.ash$loglik))
print("fitted g")
print(res.ash$fitted_g)
plot(res.ash,xmin=0,xmax=1e-4)
abline(v=res.ash$fitted_g$a,col=1)
abline(v=res.ash$fitted_g$b,col=2)
```


## `ebpm_point_gamma`:
```{r warning=F}
res.ebpm.point = ebpm::ebpm_point_gamma(d$x,d$s)
```

```{r echo = F}
print(sprintf("loglikelihood:  %f", res.ebpm.point$log_likelihood))
print("fitted g")
print(res.ebpm.point$fitted_g)
print(sprintf("mean: a/b = %5e", res.ebpm.point$fitted_g$a/res.ebpm.point$fitted_g$b))
print(sprintf("var a/b^2 =  %5e", res.ebpm.point$fitted_g$a/(res.ebpm.point$fitted_g$b^2)))

compute_cdf_point <- function(pi, a, b, max, min = 0, bins = 100){
  lams = seq(min, max, (max-min)/bins)
  cdf = pi*(as.integer(lams == 0)) + (1-pi)*pgamma(q = lams, shape = a, rate = b, lower.tail = T)
  out = list(cdf = cdf, lam = lams)
  return(out)
}

out = compute_cdf_point(res.ebpm.point$fitted_g$pi, res.ebpm.point$fitted_g$a, res.ebpm.point$fitted_g$b, max = 1e-4, bins = 1000)
plot(out$lam, out$cdf, type = "l", xlab = "lambda", ylab = "cdf")
```

## `ebpm_exponential_mixture`:
```{r warning=F}
res.ebpm.mixture = ebpm::ebpm_exponential_mixture(d$x,d$s, m = 1.1)
```

Below I  show the cdf plots, with exponential means (selected grids) of the exponentials. 
```{r, echo=F}
print(sprintf("loglikelihood:  %f", res.ebpm.mixture$log_likelihood))
print("fitted g")
print(res.ebpm.mixture$fitted_g)
idx = which.max(res.ebpm.mixture$fitted_g$pi)
print(sprintf("max pi = %f", res.ebpm.mixture$fitted_g$pi[idx]))
print(sprintf("mean for that exponential %5e", 1/res.ebpm.mixture$fitted_g$b[idx]))
print(sprintf("variance for that exponential %5e", 1/(res.ebpm.mixture$fitted_g$b[idx]^2)))


compute_cdf_mix <- function(pi, a, b, max, min = 0, bins = 100){
  lams = seq(min, max, (max-min)/bins)
  n = length(lams)
  d = length(a)
  ## P(X < x) = \sum_k  \pi_k Gamma(X < x; a_k , b_k)
  cdf = replicate(n, 0)
  ## need to speed up
  for(i in 1:n){
    for(k in 1:d){
      cdf[i] = cdf[i] + pi[k]*pgamma(q = lams[i], shape = a[k], rate = b[k], lower.tail = T)
    }
  }
  out = list(cdf = cdf, lam = lams)
  #plot(out$lam, out$cdf, type = "l")
  return(out)
}

out = compute_cdf_mix(res.ebpm.mixture$fitted_g$pi, res.ebpm.mixture$fitted_g$a, res.ebpm.mixture$fitted_g$b, max = 1e-4, bins = 1000)
plot(out$lam, out$cdf, type = "l", xlab = "lambda", ylab = "cdf")
abline(v=1/res.ebpm.mixture$fitted_g$b,col=2)
```

