---
title: "softmax_experiments"
author: "zihao12"
date: "2019-10-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Goal
I need to compute softmax along an axis in 3d array (see equation below), but it often takes too long for a moderately sized array.  

$$
\begin{align}
  \frac{exp(X_{ijk})}{\sum_k exp(X_{ijk})}
\end{align}
$$

```{r}
## x is 3d array, and we want to do softmax along the 3rd axis, so sum_k prob_ijk = 1
softmax <- function(x){
  score.exp <- exp(x)
  probs <-sweep(score.exp, MARGIN = c(1,2), apply(score.exp, MARGIN = c(1,2), sum), '/')
  return(probs)
}

# n = 1000
# p = 5000
# k = 10
# x = array(runif(n*p*k), dim = c(n, p,k)) ## this step also takes a long time!!
# start = proc.time()
# x.softmax = softmax(x)
# runtime = proc.time() - start
# print(runtime[[3]])
## [1] 12.928
# x.softmax.sum =  apply(x.softmax, MARGIN = c(1,2), sum)
# print(all.equal(x.softmax.sum, matrix(replicate(n*p, 1), nrow = n)))
## TRUE
```

Here is how scipy implements it:  ```np.exp(x - logsumexp(x, axis=axis, keepdims=True))```

So do that in the log space first:

$$
\begin{align}
  exp(X_{ijk} - \sum_k logsumexp(X_{ijk}))
\end{align}
$$

```{r}
library(matrixStats)
softmax2 <- function(x){
  k = dim(x)[3]
  x.logsumexp = apply(x, c(1,2), logSumExp)
  x.softmax = exp(x - replicate(k, x.logsumexp))
  return(x.softmax)
}
# n = 1000
# p = 5000
# k = 10
# x = array(runif(n*p*k), dim = c(n, p,k)) ## this step also takes a long time!!
# start = proc.time()
# x.softmax = softmax2(x)
# runtime = proc.time() - start
# print(runtime[[3]])
## [1] 16.445
# x.softmax.sum =  apply(x.softmax, MARGIN = c(1,2), sum)
# print(all.equal(x.softmax.sum, matrix(replicate(n*p, 1), nrow = n)))
## [1] TRUE
```
It is even slower


From Matthew:
```{r}
softmax <- function(x){
  score.exp <- exp(x)
  probs <-sweep(score.exp, MARGIN = c(1,2), apply(score.exp, MARGIN = c(1,2), sum), '/')
  return(probs)
}
softmax2 <- function(x){
  score.exp <- exp(x)
  probs <-sweep(score.exp, MARGIN = c(1,2), rowSums(score.exp,dims=2), '/')
  return(probs)
}
softmax3 <- function(x){
  score.exp <- exp(x)
  browser()
  probs <-as.vector(score.exp)/as.vector(rowSums(score.exp,dims=2))
  dim(probs) <- dim(x)
  return(probs)
}
# n = 1000
# p = 5000
# k = 10
# x = array(runif(n*p*k), dim = c(n, p,k)) ## this step also takes a long time!!
# start = proc.time()
# x.softmax = softmax(x)
# runtime = proc.time() - start
# print(runtime[[3]])
# start = proc.time()
# x.softmax2 = softmax2(x)
# runtime = proc.time() - start
# print(runtime[[3]])
# start = proc.time()
# x.softmax3 = softmax3(x)
# runtime = proc.time() - start
# print(runtime[[3]])
# identical(x.softmax3,x.softmax)
# identical(x.softmax2,x.softmax)
# 
# # [1] 11.188
# # [1] 2.662
# # [1] 1.685
# # [1] TRUE
# # [1] TRUE
```


## Lesson
* try to avoid "apply": `rowSums(score.exp,dims=2)` is much faster than apply(score.exp, MARGIN = c(1,2), sum)\
* R is column major, so be aware of what `as.vector` does \
* operations between vectors of different lengths: shorter vectors are recycled until it matches the lengths of the longer vector



