---
title: "vignette_ebpm"
author: "zihao12"
date: "2019-10-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## EBPM problem
$$
\begin{align}
  & x_i \sim Pois(s_i \lambda_i)\\
  & \lambda_i \sim g(.)\\
  & g \in \mathcal{G}
\end{align}
$$
Our goal is to estimate $\hat{g}$ (MLE), then  compute posterior $p(\lambda_i | x_i, \hat{g})$. Here  I use  mixture of exponential as prior family. 

See model details and derivations in https://github.com/stephenslab/ebpm/blob/master/derivations/ebpm.pdf

```{r warning=F}
library(ebpm)
set.seed(123)
library(ebpm)
library(ggplot2)
library(gtools)
require(gridExtra)
```


## experiment setup
I simulate data from the mixture of exponential, and compare fitting  the poisson mean problem with MLE  ($\hat{\lambda}^{mle}_i = \frac{x_i}{s_i}$), and  `ebpm_exponential_mixture` with  different options. \

For `ebpm`, the options are: \

* `fit_true_g`: use the true $g$ \

* `fit_true_scale`: use the true mixture components to estimate $\hat{g}$\

* `fit_est_scale`: estimate mixture components from data, then  estimate $\hat{g}$ 


```{r}
## simulate data
n = 2000
sim = ebpm::simulate_pois_expmix(n, seed = 123)
hist(sim$x, breaks = 100, xlab = "x", main = "hist for data x")
```


```{r}
rmse <- function(x,y){
  return(sqrt(mean((x-y)^2)))
}
methods = c()
ll_gs = c()
rmses = c()
## true
methods = c(methods, "true")
ll_gs = c(ll_gs, sim$ll)
rmses = c(rmses, 0)

## MLE
methods = c(methods, "mle")
ll_gs = c(ll_gs, NA)
rmses = c(rmses, rmse(sim$x/sim$s, sim$lam))


## fit (with known g)
fit = ebpm::ebpm_exponential_mixture(x = sim$x, s = sim$s, g_init = sim$g, fix_g = T)
methods = c(methods, "fit_true_g")
ll_gs = c(ll_gs, fit$log_likelihood)
rmses = c(rmses, rmse(fit$posterior$mean, sim$lam))
rm(fit)

## fit (with known true scale (mixture components))
fit = ebpm::ebpm_exponential_mixture(x = sim$x, s = sim$s, scale = list(shape = sim$g$shape, scale = sim$g$scale))
methods = c(methods, "fit_true_scale")
ll_gs = c(ll_gs, fit$log_likelihood)
rmses = c(rmses, rmse(fit$posterior$mean, sim$lam))
rm(fit)

## fit (estimate scale)
fit = ebpm::ebpm_exponential_mixture(x = sim$x, s = sim$s, scale = "estimate")
methods = c(methods, "fit_est_scale")
ll_gs = c(ll_gs, fit$log_likelihood)
rmses = c(rmses, rmse(fit$posterior$mean, sim$lam))
rm(fit)

data.frame(method = methods, ll_g = ll_gs, rmse = rmses)
```



## plot the $\lambda$s 
```{r}
fit = ebpm::ebpm_exponential_mixture(x = sim$x, s = sim$s, scale = "estimate")
df = data.frame(lam_true = sim$lam,lam_hat_mle = sim$x/sim$s, lam_hat_ebpm = fit$posterior$mean)
ggplot(df)+
  geom_point(aes(x = log(lam_true + 1), y = log(lam_hat_ebpm +1)), color = "blue")+
  geom_abline(slope = 1, intercept = 0)+
  guides(fill = "color")+
  ggtitle("lam_true vs lam_hat_ebpm")

ggplot(df)+
  geom_point(aes(x = log(lam_true + 1), y = log(lam_hat_mle + 1)), color = "blue")+
  geom_abline(slope = 1, intercept = 0)+
  guides(fill = "color")+
  ggtitle("lam_true vs lam_hat_mle")

ggplot(df)+
  geom_point(aes(x = log(lam_hat_mle + 1), y = log(lam_hat_ebpm + 1)), color = "blue")+
  geom_abline(slope = 1, intercept = 0)+
  guides(fill = "color")+
  ggtitle("lam_hat_mle vs lam_hat_ebpm")
```



